---
title: "Bulk RNA-seq DESeq2 Analysis — siSLC7A5 (Liver)"
author: "Jaehyun Choi"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 8,
  fig.height = 6
)
```

---

# 1. Library

```{r libraries}
library(DESeq2)
library(tidyverse)
library(pheatmap)
library(RColorBrewer)
library(ggplot2)
library(ggrepel)
library(clusterProfiler)
library(msigdbr)
library(org.Hs.eg.db)
library(EnhancedVolcano)

# ensembldb 등 다른 패키지와의 namespace 충돌 방지
filter  <- dplyr::filter
select  <- dplyr::select
arrange <- dplyr::arrange
```

---

# 2. Data Import

```{r import}
setwd("~/Liver_bio_project/02_SLC7A5_mono_Journal/04_bulkRNAseq")

# salmon length-scaled counts 사용 (DESeq2 권장)
raw <- read.table(
  "salmon.merged.gene_counts_length_scaled.tsv",
  header    = TRUE,
  row.names = 1,
  sep       = "\t",
  check.names = FALSE
)

# gene_name 컬럼 제거 → 숫자 matrix
counts_mat <- raw[, -1]

# salmon 비정수 카운트 → 반올림
counts_mat <- round(counts_mat)

cat("Dimensions:", nrow(counts_mat), "genes x", ncol(counts_mat), "samples\n")
head(counts_mat)
```

---

# 3. Sample Metadata

```{r metadata}
sample_info <- data.frame(
  sample = colnames(counts_mat),
  group  = case_when(
    grepl("^siCon_[0-9]",  colnames(counts_mat)) ~ "siCon",
    grepl("^siCon_LPS",    colnames(counts_mat)) ~ "siCon_LPS",
    grepl("^siSLC_LPS5",   colnames(counts_mat)) ~ "siSLC_LPS5",
    TRUE ~ NA_character_
  ),
  row.names = colnames(counts_mat)
)

# Reference level: siCon
sample_info$group <- factor(sample_info$group,
                            levels = c("siCon", "siCon_LPS", "siSLC_LPS5"))

print(sample_info)
table(sample_info$group)
```

---

# 4. QC — Library Size

```{r qc_libsize}
lib_df <- data.frame(
  sample       = colnames(counts_mat),
  library_size = colSums(counts_mat),
  group        = sample_info$group
)

ggplot(lib_df, aes(x = reorder(sample, library_size), y = library_size / 1e6, fill = group)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Library Size per Sample",
       x = "Sample", y = "Library size (Millions)") +
  theme_bw()
```

---

# 5. DESeq2 Dataset Construction

```{r deseq2_construct}
# Low-count gene pre-filtering (rowSum >= 10)
keep <- rowSums(counts_mat) >= 10
counts_filt <- counts_mat[keep, ]
cat("Genes after pre-filtering:", nrow(counts_filt), "\n")

dds <- DESeqDataSetFromMatrix(
  countData = counts_filt,
  colData   = sample_info,
  design    = ~ group
)

# DESeq2 실행
dds <- DESeq(dds)
dds
```

---

# 6. VST Normalization & PCA

```{r vst_pca}
vst <- vst(dds, blind = TRUE)

# PCA
pca_data <- plotPCA(vst, intgroup = "group", returnData = TRUE)
pct_var  <- round(100 * attr(pca_data, "percentVar"))

ggplot(pca_data, aes(x = PC1, y = PC2, color = group, label = name)) +
  geom_point(size = 4) +
  geom_text_repel(size = 3) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "PCA — VST normalized",
    x = paste0("PC1: ", pct_var[1], "% variance"),
    y = paste0("PC2: ", pct_var[2], "% variance")
  ) +
  theme_bw()
```

---

# 7. Sample Distance Heatmap

```{r sample_dist}
vst_mat  <- assay(vst)
samp_dist <- dist(t(vst_mat))
samp_mat  <- as.matrix(samp_dist)

ann_col <- data.frame(group = sample_info$group)
rownames(ann_col) <- rownames(sample_info)

pheatmap(
  samp_mat,
  clustering_distance_rows = samp_dist,
  clustering_distance_cols = samp_dist,
  annotation_col = ann_col,
  col = colorRampPalette(rev(brewer.pal(9, "Blues")))(255),
  main = "Sample-to-Sample Distance (VST)"
)
```

---

# 8. DEG Analysis

## 8-1. Helper function

```{r deg_helper}
get_results <- function(dds, contrast, alpha = 0.05) {
  res <- results(dds,
                 contrast  = contrast,
                 alpha     = alpha,
                 pAdjustMethod = "BH")
  res_df <- as.data.frame(res) %>%
    rownames_to_column("gene") %>%
    arrange(padj) %>%
    filter(!is.na(padj))
  return(res_df)
}
```

---

## 8-2. Comparison 1: siCon_LPS vs siCon (LPS effect)

```{r comp1}
res_LPS <- get_results(dds, contrast = c("group", "siCon_LPS", "siCon"))

cat("[siCon_LPS vs siCon]\n")
cat("  UP  (padj<0.05, log2FC>1):", sum(res_LPS$padj < 0.05 & res_LPS$log2FoldChange >  1, na.rm=TRUE), "\n")
cat("  DOWN(padj<0.05, log2FC<-1):", sum(res_LPS$padj < 0.05 & res_LPS$log2FoldChange < -1, na.rm=TRUE), "\n")

head(res_LPS, 10)
write.csv(res_LPS, "DEG_siConLPS_vs_siCon.csv", row.names = FALSE)
```

### Volcano plot

```{r volcano_comp1, fig.width=9, fig.height=7}
EnhancedVolcano(
  res_LPS,
  lab            = res_LPS$gene,
  x              = "log2FoldChange",
  y              = "padj",
  pCutoff        = 0.05,
  FCcutoff       = 1,
  title          = "siCon_LPS vs siCon",
  subtitle       = "LPS effect",
  legendPosition = "right",
  drawConnectors = TRUE
)
```

### MA plot

```{r ma_comp1}
DESeq2::plotMA(results(dds, contrast = c("group", "siCon_LPS", "siCon")),
               alpha = 0.05, main = "MA plot: siCon_LPS vs siCon")
```

---

## 8-3. Comparison 2: siSLC_LPS5 vs siCon_LPS (SLC7A5 KD effect under LPS)

```{r comp2}
res_KD <- get_results(dds, contrast = c("group", "siSLC_LPS5", "siCon_LPS"))

cat("[siSLC_LPS5 vs siCon_LPS]\n")
cat("  UP  (padj<0.05, log2FC>1):", sum(res_KD$padj < 0.05 & res_KD$log2FoldChange >  1, na.rm=TRUE), "\n")
cat("  DOWN(padj<0.05, log2FC<-1):", sum(res_KD$padj < 0.05 & res_KD$log2FoldChange < -1, na.rm=TRUE), "\n")

head(res_KD, 10)
write.csv(res_KD, "DEG_siSLCLPS5_vs_siConLPS.csv", row.names = FALSE)
```

### Volcano plot

```{r volcano_comp2, fig.width=9, fig.height=7}
EnhancedVolcano(
  res_KD,
  lab            = res_KD$gene,
  x              = "log2FoldChange",
  y              = "padj",
  pCutoff        = 0.05,
  FCcutoff       = 1,
  title          = "siSLC_LPS5 vs siCon_LPS",
  subtitle       = "SLC7A5 KD effect (LPS background)",
  legendPosition = "right",
  drawConnectors = TRUE
)
```

### MA plot

```{r ma_comp2}
DESeq2::plotMA(results(dds, contrast = c("group", "siSLC_LPS5", "siCon_LPS")),
               alpha = 0.05, main = "MA plot: siSLC_LPS5 vs siCon_LPS")
```

---

## 8-4. Comparison 3: siSLC_LPS5 vs siCon (overall effect)

```{r comp3}
res_overall <- get_results(dds, contrast = c("group", "siSLC_LPS5", "siCon"))

cat("[siSLC_LPS5 vs siCon]\n")
cat("  UP  (padj<0.05, log2FC>1):", sum(res_overall$padj < 0.05 & res_overall$log2FoldChange >  1, na.rm=TRUE), "\n")
cat("  DOWN(padj<0.05, log2FC<-1):", sum(res_overall$padj < 0.05 & res_overall$log2FoldChange < -1, na.rm=TRUE), "\n")

head(res_overall, 10)
write.csv(res_overall, "DEG_siSLCLPS5_vs_siCon.csv", row.names = FALSE)
```

### Volcano plot

```{r volcano_comp3, fig.width=9, fig.height=7}
EnhancedVolcano(
  res_overall,
  lab            = res_overall$gene,
  x              = "log2FoldChange",
  y              = "padj",
  pCutoff        = 0.05,
  FCcutoff       = 1,
  title          = "siSLC_LPS5 vs siCon",
  subtitle       = "Overall SLC7A5 KD + LPS effect",
  legendPosition = "right",
  drawConnectors = TRUE
)
```

---

# 9. Top DEG Heatmap

```{r deg_heatmap, fig.width=10, fig.height=8}
# 각 비교에서 유의미한 DEG 통합 (padj<0.05, |log2FC|>1)
sig_genes <- unique(c(
  res_LPS     %>% filter(padj < 0.05, abs(log2FoldChange) > 1) %>% pull(gene),
  res_KD      %>% filter(padj < 0.05, abs(log2FoldChange) > 1) %>% pull(gene),
  res_overall %>% filter(padj < 0.05, abs(log2FoldChange) > 1) %>% pull(gene)
))

cat("Total significant DEGs (union):", length(sig_genes), "\n")

# VST matrix에서 해당 유전자만 추출
vst_sig <- vst_mat[rownames(vst_mat) %in% sig_genes, ]

# 최대 100개 유전자 (variance 기준)
if (nrow(vst_sig) > 100) {
  gene_var <- apply(vst_sig, 1, var)
  vst_sig  <- vst_sig[order(gene_var, decreasing = TRUE)[1:100], ]
}

ann_col <- data.frame(group = sample_info$group)
rownames(ann_col) <- rownames(sample_info)

group_colors <- list(group = c(
  siCon      = "#4393C3",
  siCon_LPS  = "#D6604D",
  siSLC_LPS5 = "#74C476"
))

pheatmap(
  vst_sig,
  scale              = "row",
  annotation_col     = ann_col,
  annotation_colors  = group_colors,
  show_rownames      = nrow(vst_sig) <= 50,
  clustering_method  = "ward.D2",
  color              = colorRampPalette(c("#2166AC", "white", "#B2182B"))(200),
  main               = paste0("Top DEG Heatmap (n=", nrow(vst_sig), ")")
)
```

---

# 10. SLC7A5 Expression

```{r slc7a5_expr}
tpm_raw <- read.table(
  "salmon.merged.gene_tpm.tsv",
  header = TRUE, row.names = 1, sep = "\t", check.names = FALSE
)
tpm_mat <- tpm_raw[, -1]
log_tpm <- log2(tpm_mat + 1)

if ("SLC7A5" %in% rownames(log_tpm)) {
  expr_df <- data.frame(
    sample = colnames(log_tpm),
    expr   = as.numeric(log_tpm["SLC7A5", ]),
    group  = sample_info$group
  )

  ggplot(expr_df, aes(x = group, y = expr, fill = group)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.7) +
    geom_jitter(width = 0.15, size = 3, color = "black") +
    scale_fill_brewer(palette = "Set1") +
    labs(title = "SLC7A5 Expression",
         y = "log2(TPM + 1)", x = "") +
    theme_bw() +
    theme(legend.position = "none")
} else {
  cat("SLC7A5 not found in TPM matrix\n")
}
```

---

# 11. GSEA — Hallmark Gene Sets (siSLC_LPS5 vs siCon_LPS)

```{r gsea, fig.width=10, fig.height=7}
# Ranked gene list (log2FC 기준)
ranking <- res_KD %>%
  filter(!is.na(log2FoldChange)) %>%
  arrange(desc(log2FoldChange)) %>%
  { setNames(.$log2FoldChange, .$gene) }

# MSigDB Hallmark
msig_h <- msigdbr(species = "Homo sapiens", category = "H") %>%
  dplyr::select(gs_name, gene_symbol)

gsea_res <- GSEA(
  geneList    = ranking,
  TERM2GENE   = msig_h,
  pvalueCutoff = 0.05,
  pAdjustMethod = "BH",
  verbose     = FALSE
)

if (nrow(gsea_res@result) > 0) {
  dotplot(gsea_res, showCategory = 20, title = "GSEA Hallmark: siSLC_LPS5 vs siCon_LPS") +
    theme(axis.text.y = element_text(size = 9))
} else {
  cat("No significant GSEA results (padj < 0.05)\n")
}
```

```{r gsea_ridge, fig.width=10, fig.height=7}
if (nrow(gsea_res@result) > 0) {
  ridgeplot(gsea_res, showCategory = 15) +
    labs(title = "GSEA Ridgeplot: siSLC_LPS5 vs siCon_LPS")
}
```

---

# 12. ORA — GO & KEGG (siSLC_LPS5 vs siCon_LPS)

```{r ora}
# 유의미하게 올라간 유전자
up_genes <- res_KD %>%
  filter(padj < 0.05, log2FoldChange > 1) %>%
  pull(gene)

dn_genes <- res_KD %>%
  filter(padj < 0.05, log2FoldChange < -1) %>%
  pull(gene)

cat("UP genes:", length(up_genes), " | DOWN genes:", length(dn_genes), "\n")
```

### GO Biological Process — UP

```{r go_bp_up, fig.width=10, fig.height=7}
if (length(up_genes) > 5) {
  go_up <- enrichGO(
    gene          = up_genes,
    OrgDb         = org.Hs.eg.db,
    keyType       = "SYMBOL",
    ont           = "BP",
    pAdjustMethod = "BH",
    pvalueCutoff  = 0.05,
    qvalueCutoff  = 0.2
  )
  if (nrow(go_up@result) > 0) {
    dotplot(go_up, showCategory = 20,
            title = "GO BP — UP in siSLC_LPS5 vs siCon_LPS")
  } else {
    cat("No significant GO BP results for UP genes\n")
  }
} else {
  cat("Not enough UP genes for ORA\n")
}
```

### GO Biological Process — DOWN

```{r go_bp_dn, fig.width=10, fig.height=7}
if (length(dn_genes) > 5) {
  go_dn <- enrichGO(
    gene          = dn_genes,
    OrgDb         = org.Hs.eg.db,
    keyType       = "SYMBOL",
    ont           = "BP",
    pAdjustMethod = "BH",
    pvalueCutoff  = 0.05,
    qvalueCutoff  = 0.2
  )
  if (nrow(go_dn@result) > 0) {
    dotplot(go_dn, showCategory = 20,
            title = "GO BP — DOWN in siSLC_LPS5 vs siCon_LPS")
  } else {
    cat("No significant GO BP results for DOWN genes\n")
  }
} else {
  cat("Not enough DOWN genes for ORA\n")
}
```

### KEGG Pathway — UP

```{r kegg_up, fig.width=10, fig.height=6}
if (length(up_genes) > 5) {
  # SYMBOL → ENTREZID 변환
  eg_up <- bitr(up_genes, fromType = "SYMBOL",
                toType = "ENTREZID", OrgDb = org.Hs.eg.db)

  kegg_up <- enrichKEGG(
    gene          = eg_up$ENTREZID,
    organism      = "hsa",
    pAdjustMethod = "BH",
    pvalueCutoff  = 0.05
  )
  if (!is.null(kegg_up) && nrow(kegg_up@result) > 0) {
    dotplot(kegg_up, showCategory = 20,
            title = "KEGG — UP in siSLC_LPS5 vs siCon_LPS")
  } else {
    cat("No significant KEGG results for UP genes\n")
  }
}
```

---

# 13. SLC Family Gene Analysis: siCon_LPS vs siSLC_LPS5 (Fisher's Exact Test)

## 13-1. SLC family DEG 추출

```{r slc_extract}
# res_KD = siSLC_LPS5 vs siCon_LPS 결과 사용
slc_res <- res_KD %>%
  dplyr::filter(grepl("^SLC", gene)) %>%
  dplyr::arrange(padj)

cat("Total SLC family genes in results:", nrow(slc_res), "\n")
cat("SLC DEGs (padj<0.05, |log2FC|>1):",
    sum(slc_res$padj < 0.05 & abs(slc_res$log2FoldChange) > 1, na.rm = TRUE), "\n")
cat("  UP  (log2FC > 1):",
    sum(slc_res$padj < 0.05 & slc_res$log2FoldChange >  1, na.rm = TRUE), "\n")
cat("  DOWN(log2FC < -1):",
    sum(slc_res$padj < 0.05 & slc_res$log2FoldChange < -1, na.rm = TRUE), "\n\n")

# 유의미한 SLC DEG 목록
slc_sig <- slc_res %>%
  dplyr::filter(padj < 0.05, abs(log2FoldChange) > 1)

print(slc_sig)
write.csv(slc_res, "SLC_family_DEG_siSLCLPS5_vs_siConLPS.csv", row.names = FALSE)
```

---

## 13-2. Fisher's Exact Test — SLC family 유전자군 enrichment

SLC family 유전자가 DEG 목록에서 유의미하게 과대/과소 표현되는지 검정합니다.

```{r slc_fisher_enrichment}
# 전체 결과에서 분류
all_res <- res_KD %>% dplyr::filter(!is.na(padj))

n_total    <- nrow(all_res)
n_slc      <- sum(grepl("^SLC", all_res$gene))
n_nonslc   <- n_total - n_slc

fisher_results <- list()

for (direction in c("UP", "DOWN")) {
  if (direction == "UP") {
    is_deg <- all_res$padj < 0.05 & all_res$log2FoldChange >  1
  } else {
    is_deg <- all_res$padj < 0.05 & all_res$log2FoldChange < -1
  }

  is_slc <- grepl("^SLC", all_res$gene)

  # 2x2 contingency table
  #              | DEG | non-DEG |
  # SLC family   |  a  |    b    |
  # non-SLC      |  c  |    d    |
  a <- sum( is_slc &  is_deg, na.rm = TRUE)
  b <- sum( is_slc & !is_deg, na.rm = TRUE)
  c <- sum(!is_slc &  is_deg, na.rm = TRUE)
  d <- sum(!is_slc & !is_deg, na.rm = TRUE)

  ct <- matrix(c(a, c, b, d), nrow = 2,
               dimnames = list(c("SLC", "non-SLC"), c("DEG", "non-DEG")))

  ft <- fisher.test(ct)

  fisher_results[[direction]] <- data.frame(
    Direction    = direction,
    SLC_DEG      = a,
    SLC_nonDEG   = b,
    nonSLC_DEG   = c,
    nonSLC_nonDEG= d,
    OddsRatio    = round(ft$estimate, 4),
    p_value      = ft$p.value,
    CI_lower     = round(ft$conf.int[1], 4),
    CI_upper     = round(ft$conf.int[2], 4),
    Significant  = ft$p.value < 0.05
  )

  cat(sprintf("[SLC enrichment — %s DEGs]\n", direction))
  print(ct)
  cat(sprintf("  Odds Ratio : %.4f (95%% CI: %.4f – %.4f)\n",
              ft$estimate, ft$conf.int[1], ft$conf.int[2]))
  cat(sprintf("  p-value    : %.4e  %s\n\n",
              ft$p.value,
              ifelse(ft$p.value < 0.001, "***",
                     ifelse(ft$p.value < 0.01, "**",
                            ifelse(ft$p.value < 0.05, "*", "ns")))))
}

enrich_df <- do.call(rbind, fisher_results)
rownames(enrich_df) <- NULL
print(enrich_df)
```

---

## 13-3. Per-gene Fisher's Exact Test — SLC 유전자별 개별 검정

각 SLC 유전자에 대해 raw count 기반 2×2 Fisher's exact test를 수행합니다.

```{r slc_per_gene_fisher}
# siCon_LPS, siSLC_LPS5 샘플 컬럼 추출
ctrl_cols <- grep("siCon_LPS",  colnames(counts_filt), value = TRUE)
kd_cols   <- grep("siSLC_LPS5", colnames(counts_filt), value = TRUE)

cat("siCon_LPS samples :", paste(ctrl_cols, collapse = ", "), "\n")
cat("siSLC_LPS5 samples:", paste(kd_cols,   collapse = ", "), "\n\n")

# 라이브러리 크기 (그룹 합계)
lib_ctrl <- sum(counts_filt[, ctrl_cols])
lib_kd   <- sum(counts_filt[, kd_cols])

# SLC 유전자만 추출
slc_genes_in_counts <- rownames(counts_filt)[grepl("^SLC", rownames(counts_filt))]
slc_counts <- counts_filt[slc_genes_in_counts, c(ctrl_cols, kd_cols)]

# Per-gene Fisher's exact test
per_gene_fisher <- lapply(slc_genes_in_counts, function(gene) {
  g_ctrl <- sum(slc_counts[gene, ctrl_cols])
  g_kd   <- sum(slc_counts[gene, kd_cols])

  # 2x2 table
  #            | this gene | other genes  |
  # siCon_LPS  |  g_ctrl   | lib_ctrl - g_ctrl |
  # siSLC_LPS5 |  g_kd     | lib_kd - g_kd     |
  ct <- matrix(
    c(g_ctrl, g_kd,
      lib_ctrl - g_ctrl, lib_kd - g_kd),
    nrow = 2,
    dimnames = list(c("siCon_LPS", "siSLC_LPS5"), c("this_gene", "other_genes"))
  )

  ft <- tryCatch(fisher.test(ct), error = function(e) NULL)
  if (is.null(ft)) return(NULL)

  data.frame(
    gene         = gene,
    count_ctrl   = g_ctrl,
    count_kd     = g_kd,
    OddsRatio    = round(ft$estimate, 4),
    p_value      = ft$p.value,
    stringsAsFactors = FALSE
  )
})

per_gene_df <- do.call(rbind, Filter(Negate(is.null), per_gene_fisher))

# BH correction
per_gene_df$p_adj <- p.adjust(per_gene_df$p_value, method = "BH")
per_gene_df$log2OR <- log2(per_gene_df$OddsRatio + 1e-6)

# DESeq2 결과와 합치기
per_gene_df <- per_gene_df %>%
  dplyr::left_join(
    slc_res %>% dplyr::select(gene, log2FoldChange, padj),
    by = "gene"
  ) %>%
  dplyr::arrange(p_adj)

cat("SLC genes significant by Fisher (p_adj < 0.05):",
    sum(per_gene_df$p_adj < 0.05, na.rm = TRUE), "\n\n")

print(head(per_gene_df, 20))
write.csv(per_gene_df, "SLC_per_gene_fisher_siSLCLPS5_vs_siConLPS.csv", row.names = FALSE)
```

---

## 13-4. Visualization

### Volcano plot — SLC genes 강조

```{r slc_volcano, fig.width=10, fig.height=7}
# res_KD에 SLC 여부 표시
plot_df <- res_KD %>%
  dplyr::mutate(
    is_slc = grepl("^SLC", gene),
    sig    = !is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 1,
    label  = ifelse(is_slc & sig, gene, NA)
  )

ggplot(plot_df, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(data = dplyr::filter(plot_df, !is_slc),
             color = "grey80", size = 0.8, alpha = 0.5) +
  geom_point(data = dplyr::filter(plot_df,  is_slc & !sig),
             color = "#90CAF9", size = 1.5, alpha = 0.8) +
  geom_point(data = dplyr::filter(plot_df,  is_slc &  sig),
             aes(color = log2FoldChange > 0), size = 2.5) +
  scale_color_manual(values = c("TRUE" = "#D32F2F", "FALSE" = "#1565C0"),
                     labels = c("TRUE" = "SLC UP", "FALSE" = "SLC DOWN")) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = c(-1, 1),     linetype = "dashed", color = "grey40") +
  geom_text_repel(aes(label = label), size = 3, max.overlaps = 20,
                  segment.color = "grey50") +
  labs(
    title    = "Volcano Plot: siSLC_LPS5 vs siCon_LPS",
    subtitle = "SLC family genes highlighted",
    x        = "log2 Fold Change",
    y        = "-log10(padj)",
    color    = ""
  ) +
  theme_bw() +
  theme(legend.position = "top")
```

### SLC DEG log2FC bar chart

```{r slc_bar, fig.width=10, fig.height=6}
slc_sig_bar <- per_gene_df %>%
  dplyr::filter(p_adj < 0.05) %>%
  dplyr::arrange(log2FoldChange)

if (nrow(slc_sig_bar) > 0) {
  slc_sig_bar$gene <- factor(slc_sig_bar$gene, levels = slc_sig_bar$gene)

  ggplot(slc_sig_bar, aes(x = gene, y = log2FoldChange,
                           fill = log2FoldChange > 0)) +
    geom_bar(stat = "identity") +
    geom_text(aes(label = sprintf("p=%.3f", p_adj),
                  y = ifelse(log2FoldChange > 0,
                             log2FoldChange + 0.05,
                             log2FoldChange - 0.05)),
              hjust = ifelse(slc_sig_bar$log2FoldChange > 0, 0, 1),
              size = 3) +
    coord_flip() +
    scale_fill_manual(values = c("TRUE" = "#D32F2F", "FALSE" = "#1565C0"),
                      labels = c("TRUE" = "UP", "FALSE" = "DOWN")) +
    labs(
      title = "Significant SLC Family Genes (Fisher p_adj < 0.05)",
      subtitle = "siSLC_LPS5 vs siCon_LPS",
      x = "", y = "log2 Fold Change (DESeq2)", fill = ""
    ) +
    theme_bw() +
    theme(legend.position = "top")
} else {
  cat("No SLC genes significant after BH correction (p_adj < 0.05)\n")
  # p_adj 기준 완화하여 top 20 표시
  top20_slc <- per_gene_df %>% dplyr::slice_head(n = 20)
  top20_slc$gene <- factor(top20_slc$gene,
                            levels = top20_slc$gene[order(top20_slc$log2FoldChange)])

  ggplot(top20_slc, aes(x = gene, y = log2FoldChange, fill = log2FoldChange > 0)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    scale_fill_manual(values = c("TRUE" = "#D32F2F", "FALSE" = "#1565C0"),
                      labels = c("TRUE" = "UP", "FALSE" = "DOWN")) +
    labs(title = "Top 20 SLC Genes by Fisher p-value (uncorrected)",
         x = "", y = "log2 Fold Change", fill = "") +
    theme_bw()
}
```

### SLC gene expression heatmap

```{r slc_heatmap, fig.width=10, fig.height=8}
# VST matrix에서 SLC 유전자 추출
slc_in_vst <- rownames(vst_mat)[grepl("^SLC", rownames(vst_mat))]

# Fisher p_adj < 0.05 또는 top 30 (p_value 기준)
slc_show <- per_gene_df %>%
  { if (sum(.$p_adj < 0.05) >= 5)
      dplyr::filter(., p_adj < 0.05)
    else
      dplyr::slice_head(., n = 30) } %>%
  dplyr::pull(gene) %>%
  intersect(slc_in_vst)

if (length(slc_show) >= 2) {
  vst_slc <- vst_mat[slc_show, c(ctrl_cols, kd_cols)]

  ann_col_slc <- data.frame(
    group = c(rep("siCon_LPS", length(ctrl_cols)),
              rep("siSLC_LPS5", length(kd_cols)))
  )
  rownames(ann_col_slc) <- c(ctrl_cols, kd_cols)

  group_colors_slc <- list(group = c(
    siCon_LPS  = "#D6604D",
    siSLC_LPS5 = "#74C476"
  ))

  pheatmap(
    vst_slc,
    scale             = "row",
    annotation_col    = ann_col_slc,
    annotation_colors = group_colors_slc,
    clustering_method = "ward.D2",
    color             = colorRampPalette(c("#2166AC", "white", "#B2182B"))(200),
    show_rownames     = TRUE,
    fontsize_row      = 8,
    main              = "SLC Family Gene Expression Heatmap\n(siCon_LPS vs siSLC_LPS5)"
  )
} else {
  cat("Not enough SLC genes to draw heatmap\n")
}
```

---

# 14. Session Info

```{r session_info}
sessionInfo()
```
