---
title: "Bulk RNA-seq DESeq2 Analysis - Additional expression files"
author: "Jaehyun Choi"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r repository-setup, include=FALSE}
.slc_start <- Sys.getenv("SLC7A5_REPO_ROOT", unset = getwd())
.slc_repo <- normalizePath(.slc_start, mustWork = TRUE)
while (!file.exists(file.path(.slc_repo, "R", "paths.R"))) {
  .slc_parent <- dirname(.slc_repo)
  if (identical(.slc_parent, .slc_repo)) stop("Start inside the repository or set SLC7A5_REPO_ROOT.")
  .slc_repo <- .slc_parent
}
source(file.path(.slc_repo, "R", "paths.R"))
slc_setup("bulk_additional")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 8,
  fig.height = 6
)
```

# 1. Library

```{r libraries}
library(DESeq2)
library(tidyverse)
library(readxl)
library(pheatmap)
library(RColorBrewer)
library(ggplot2)
library(ggrepel)
library(clusterProfiler)
library(msigdbr)
library(org.Hs.eg.db)
library(EnhancedVolcano)

filter  <- dplyr::filter
select  <- dplyr::select
arrange <- dplyr::arrange
```

# 2. Input Files

```{r paths}
additional_dir <- slc_output_dir("02_SLC7A5_mono_Journal/04_bulkRNAseq/additional")
project_dir <- slc_output_dir("02_SLC7A5_mono_Journal/04_bulkRNAseq")
gene_expression_file <- slc_input("bulk_additional_genes")
transcript_expression_file <- slc_input("bulk_additional_transcripts")

stopifnot(file.exists(gene_expression_file))
stopifnot(file.exists(transcript_expression_file))

output_suffix <- "additional"

# Keep all samples in this additional dataset, including Cont1.
exclude_baseline_sample <- NULL
```

# 3. Data Import

```{r import_helpers}
read_expression_xlsx <- function(path, id_col, feature_label = "feature",
                                 exclude_sample = NULL) {
  raw <- readxl::read_excel(path, sheet = 1) %>%
    as.data.frame()

  count_cols <- grep("^EXP:.*:ReadCount$", colnames(raw), value = TRUE)
  tpm_cols   <- grep("^EXP:.*:TPM$",       colnames(raw), value = TRUE)

  if (length(count_cols) == 0) {
    stop("No ReadCount columns found in: ", path)
  }
  if (length(tpm_cols) == 0) {
    stop("No TPM columns found in: ", path)
  }

  count_samples <- sub("^EXP:(.*):ReadCount$", "\\1", count_cols)
  tpm_samples   <- sub("^EXP:(.*):TPM$",       "\\1", tpm_cols)

  counts_mat <- raw[, count_cols, drop = FALSE]
  tpm_mat    <- raw[, tpm_cols,   drop = FALSE]

  colnames(counts_mat) <- count_samples
  colnames(tpm_mat)    <- tpm_samples

  counts_mat <- as.matrix(counts_mat)
  tpm_mat    <- as.matrix(tpm_mat)
  storage.mode(counts_mat) <- "numeric"
  storage.mode(tpm_mat)    <- "numeric"

  if (!is.null(exclude_sample)) {
    counts_mat <- counts_mat[, colnames(counts_mat) != exclude_sample, drop = FALSE]
    tpm_mat    <- tpm_mat[,    colnames(tpm_mat)    != exclude_sample, drop = FALSE]
  }

  feature_symbol <- raw$Symbol
  feature_id <- ifelse(
    !is.na(feature_symbol) & feature_symbol != "",
    feature_symbol,
    raw[[id_col]]
  )
  feature_id <- make.unique(as.character(feature_id))

  rownames(counts_mat) <- feature_id
  rownames(tpm_mat)    <- feature_id

  annotation <- raw %>%
    dplyr::transmute(
      feature = feature_id,
      source_id = .data[[id_col]],
      symbol = Symbol,
      biotype = Biotype,
      description = Description
    )

  list(
    raw = raw,
    counts = round(counts_mat),
    tpm = tpm_mat,
    annotation = annotation,
    feature_label = feature_label
  )
}

make_sample_info <- function(sample_names) {
  sample_info <- data.frame(
    sample = sample_names,
    group  = dplyr::case_when(
      grepl("^Cont[0-9]+$",   sample_names) ~ "siCon",
      grepl("^Cont_L[0-9]+$", sample_names) ~ "siCon_LPS",
      grepl("^SLC_L[0-9]+$",  sample_names) ~ "siSLC_LPS5",
      TRUE ~ NA_character_
    ),
    row.names = sample_names
  )

  if (any(is.na(sample_info$group))) {
    stop("Unmapped sample(s): ",
         paste(sample_info$sample[is.na(sample_info$group)], collapse = ", "))
  }

  sample_info$group <- factor(sample_info$group,
                              levels = c("siCon", "siCon_LPS", "siSLC_LPS5"))
  sample_info
}

gene_expr <- read_expression_xlsx(
  gene_expression_file,
  id_col = "GeneId",
  feature_label = "gene",
  exclude_sample = exclude_baseline_sample
)

tx_expr <- read_expression_xlsx(
  transcript_expression_file,
  id_col = "TranId",
  feature_label = "transcript",
  exclude_sample = exclude_baseline_sample
)

counts_mat <- gene_expr$counts
tpm_mat <- gene_expr$tpm
sample_info <- make_sample_info(colnames(counts_mat))

cat("Gene matrix:", nrow(counts_mat), "genes x", ncol(counts_mat), "samples\n")
cat("Transcript matrix:", nrow(tx_expr$counts), "transcripts x",
    ncol(tx_expr$counts), "samples\n\n")
print(sample_info)
table(sample_info$group)
```

# 4. QC - Library Size

```{r qc_libsize}
lib_df <- data.frame(
  sample       = colnames(counts_mat),
  library_size = colSums(counts_mat),
  group        = sample_info$group
)

ggplot(lib_df, aes(x = reorder(sample, library_size),
                   y = library_size / 1e6, fill = group)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Gene-level library size per sample",
       x = "Sample", y = "Library size (Millions)") +
  theme_bw()
```

# 5. DESeq2 Dataset Construction

```{r deseq_helpers}
run_deseq2 <- function(counts_mat, sample_info, min_row_sum = 10) {
  keep <- rowSums(counts_mat) >= min_row_sum
  counts_filt <- counts_mat[keep, , drop = FALSE]

  dds <- DESeqDataSetFromMatrix(
    countData = counts_filt,
    colData   = sample_info,
    design    = ~ group
  )
  dds <- DESeq(dds)

  list(dds = dds, counts_filt = counts_filt)
}

get_results <- function(dds, contrast, alpha = 0.05) {
  res <- results(dds,
                 contrast  = contrast,
                 alpha     = alpha,
                 pAdjustMethod = "BH")
  as.data.frame(res) %>%
    rownames_to_column("gene") %>%
    arrange(padj) %>%
    filter(!is.na(padj))
}

summarize_deg <- function(res_df, label, lfc_cutoff = 0.58) {
  cat("[", label, "]\n", sep = "")
  cat("  UP  (padj<0.05, log2FC>", lfc_cutoff, "): ",
      sum(res_df$padj < 0.05 & res_df$log2FoldChange > lfc_cutoff, na.rm = TRUE), "\n", sep = "")
  cat("  DOWN(padj<0.05, log2FC<", -lfc_cutoff, "): ",
      sum(res_df$padj < 0.05 & res_df$log2FoldChange < -lfc_cutoff, na.rm = TRUE), "\n\n", sep = "")
}
```

```{r deseq2_construct}
gene_de <- run_deseq2(counts_mat, sample_info)
dds <- gene_de$dds
counts_filt <- gene_de$counts_filt

cat("Genes after pre-filtering:", nrow(counts_filt), "\n")
cat("DESeq2 design:", as.character(design(dds)), "\n")
dds
```

# 6. VST Normalization & PCA

```{r vst_pca}
vst <- vst(dds, blind = TRUE)
vst_mat <- assay(vst)

pca_data <- plotPCA(vst, intgroup = "group", returnData = TRUE)
pct_var  <- round(100 * attr(pca_data, "percentVar"))

ggplot(pca_data, aes(x = PC1, y = PC2, color = group, label = name)) +
  geom_point(size = 4) +
  geom_text_repel(size = 3) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "PCA - VST normalized gene counts",
    x = paste0("PC1: ", pct_var[1], "% variance"),
    y = paste0("PC2: ", pct_var[2], "% variance")
  ) +
  theme_bw()
```

# 7. Sample Distance Heatmap

```{r sample_dist}
samp_dist <- dist(t(vst_mat))
samp_mat  <- as.matrix(samp_dist)

ann_col <- data.frame(group = sample_info$group)
rownames(ann_col) <- rownames(sample_info)

pheatmap(
  samp_mat,
  clustering_distance_rows = samp_dist,
  clustering_distance_cols = samp_dist,
  annotation_col = ann_col,
  col = colorRampPalette(rev(brewer.pal(9, "Blues")))(255),
  main = "Sample-to-Sample Distance (VST, gene-level)"
)
```

# 8. DEG Analysis - Gene Level

## 8-1. Comparison 1: siCon_LPS vs siCon

```{r comp1}
res_LPS <- get_results(dds, contrast = c("group", "siCon_LPS", "siCon"))
summarize_deg(res_LPS, "siCon_LPS vs siCon")
head(res_LPS, 10)
write.csv(res_LPS,
          file.path(additional_dir, paste0("DEG_siConLPS_vs_siCon_", output_suffix, ".csv")),
          row.names = FALSE)
```

```{r volcano_comp1, fig.width=9, fig.height=7}
EnhancedVolcano(
  res_LPS,
  lab            = res_LPS$gene,
  x              = "log2FoldChange",
  y              = "padj",
  pCutoff        = 0.05,
  FCcutoff       = 0.58,
  title          = "siCon_LPS vs siCon",
  subtitle       = "LPS effect",
  legendPosition = "right",
  drawConnectors = TRUE
)
```

```{r ma_comp1}
DESeq2::plotMA(results(dds, contrast = c("group", "siCon_LPS", "siCon")),
               alpha = 0.05, main = "MA plot: siCon_LPS vs siCon")
```

## 8-2. Comparison 2: siSLC_LPS5 vs siCon_LPS

```{r comp2}
res_KD <- get_results(dds, contrast = c("group", "siSLC_LPS5", "siCon_LPS"))
summarize_deg(res_KD, "siSLC_LPS5 vs siCon_LPS")
head(res_KD, 10)
write.csv(res_KD,
          file.path(additional_dir, paste0("DEG_siSLCLPS5_vs_siConLPS_", output_suffix, ".csv")),
          row.names = FALSE)
```

```{r volcano_comp2, fig.width=9, fig.height=7}
EnhancedVolcano(
  res_KD,
  lab            = res_KD$gene,
  x              = "log2FoldChange",
  y              = "padj",
  pCutoff        = 0.05,
  FCcutoff       = 0.58,
  title          = "siSLC_LPS5 vs siCon_LPS",
  subtitle       = "SLC7A5 KD effect (LPS background)",
  legendPosition = "right",
  drawConnectors = TRUE
)
```

```{r ma_comp2}
DESeq2::plotMA(results(dds, contrast = c("group", "siSLC_LPS5", "siCon_LPS")),
               alpha = 0.05, main = "MA plot: siSLC_LPS5 vs siCon_LPS")
```

## 8-3. Comparison 3: siSLC_LPS5 vs siCon

```{r comp3}
res_overall <- get_results(dds, contrast = c("group", "siSLC_LPS5", "siCon"))
summarize_deg(res_overall, "siSLC_LPS5 vs siCon")
head(res_overall, 10)
write.csv(res_overall,
          file.path(additional_dir, paste0("DEG_siSLCLPS5_vs_siCon_", output_suffix, ".csv")),
          row.names = FALSE)
```

```{r volcano_comp3, fig.width=9, fig.height=7}
EnhancedVolcano(
  res_overall,
  lab            = res_overall$gene,
  x              = "log2FoldChange",
  y              = "padj",
  pCutoff        = 0.05,
  FCcutoff       = 0.58,
  title          = "siSLC_LPS5 vs siCon",
  subtitle       = "Overall SLC7A5 KD + LPS effect",
  legendPosition = "right",
  drawConnectors = TRUE
)
```

# 9. Top DEG Heatmap

```{r deg_heatmap, fig.width=10, fig.height=8}
sig_genes <- unique(c(
  res_LPS     %>% filter(padj < 0.05, abs(log2FoldChange) > 0.58) %>% pull(gene),
  res_KD      %>% filter(padj < 0.05, abs(log2FoldChange) > 0.58) %>% pull(gene),
  res_overall %>% filter(padj < 0.05, abs(log2FoldChange) > 0.58) %>% pull(gene)
))

cat("Total significant DEGs (union):", length(sig_genes), "\n")

vst_sig <- vst_mat[rownames(vst_mat) %in% sig_genes, , drop = FALSE]

if (nrow(vst_sig) > 100) {
  gene_var <- apply(vst_sig, 1, var)
  vst_sig  <- vst_sig[order(gene_var, decreasing = TRUE)[1:100], , drop = FALSE]
}

group_colors <- list(group = c(
  siCon      = "#4393C3",
  siCon_LPS  = "#D6604D",
  siSLC_LPS5 = "#74C476"
))

if (nrow(vst_sig) >= 2) {
  pheatmap(
    vst_sig,
    scale              = "row",
    annotation_col     = ann_col,
    annotation_colors  = group_colors,
    show_rownames      = nrow(vst_sig) <= 50,
    clustering_method  = "ward.D2",
    color              = colorRampPalette(c("#2166AC", "white", "#B2182B"))(200),
    main               = paste0("Top DEG Heatmap (n=", nrow(vst_sig), ")")
  )
} else {
  cat("Not enough significant genes for DEG heatmap\n")
}
```

# 10. SLC7A5 Expression

```{r slc7a5_expr}
log_tpm <- log2(tpm_mat + 1)

if ("SLC7A5" %in% rownames(log_tpm)) {
  expr_df <- data.frame(
    sample = colnames(log_tpm),
    expr   = as.numeric(log_tpm["SLC7A5", ]),
    group  = sample_info$group
  )

  pairwise_comparisons <- tibble::tribble(
    ~group1,       ~group2,
    "siCon",       "siCon_LPS",
    "siCon_LPS",   "siSLC_LPS5",
    "siCon",       "siSLC_LPS5"
  ) %>%
    dplyr::filter(group1 %in% levels(expr_df$group),
                  group2 %in% levels(expr_df$group))

  stat_df <- purrr::pmap_dfr(pairwise_comparisons, function(group1, group2) {
    test_df <- expr_df %>%
      dplyr::filter(group %in% c(group1, group2)) %>%
      dplyr::mutate(group = droplevels(group))

    p_value <- tryCatch(
      t.test(expr ~ group, data = test_df)$p.value,
      error = function(e) NA_real_
    )

    tibble::tibble(group1 = group1, group2 = group2, p_value = p_value)
  }) %>%
    dplyr::mutate(
      p_adj = p.adjust(p_value, method = "BH"),
      label = dplyr::case_when(
        is.na(p_adj) ~ "NA",
        p_adj < 0.001 ~ "***",
        p_adj < 0.01  ~ "**",
        p_adj < 0.05  ~ "*",
        TRUE ~ "ns"
      )
    )

  y_range <- diff(range(expr_df$expr, na.rm = TRUE))
  y_step <- ifelse(y_range == 0, 0.2, y_range * 0.12)
  y_base <- max(expr_df$expr, na.rm = TRUE)
  group_levels <- levels(expr_df$group)

  stat_df <- stat_df %>%
    dplyr::mutate(
      x1 = match(group1, group_levels),
      x2 = match(group2, group_levels),
      y = y_base + dplyr::row_number() * y_step,
      y_text = y + y_step * 0.18
    )

  p <- ggplot(expr_df, aes(x = group, y = expr, fill = group)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.7) +
    geom_jitter(width = 0.15, size = 3, color = "black") +
    scale_fill_brewer(palette = "Set1") +
    labs(title = "SLC7A5 Expression",
         y = "log2(TPM + 1)", x = "") +
    theme_bw() +
    theme(legend.position = "none")

  if (nrow(stat_df) > 0) {
    p <- p +
      geom_segment(data = stat_df,
                   aes(x = x1, xend = x2, y = y, yend = y),
                   inherit.aes = FALSE) +
      geom_segment(data = stat_df,
                   aes(x = x1, xend = x1, y = y - y_step * 0.12, yend = y),
                   inherit.aes = FALSE) +
      geom_segment(data = stat_df,
                   aes(x = x2, xend = x2, y = y - y_step * 0.12, yend = y),
                   inherit.aes = FALSE) +
      geom_text(data = stat_df,
                aes(x = (x1 + x2) / 2, y = y_text, label = label),
                inherit.aes = FALSE, size = 5) +
      expand_limits(y = max(stat_df$y_text, na.rm = TRUE) + y_step * 0.3)

    print(stat_df %>% dplyr::select(group1, group2, p_value, p_adj, label))
  }

  p
} else {
  cat("SLC7A5 not found in TPM matrix\n")
}
```

# 11. GSEA - Hallmark Gene Sets

```{r gsea, fig.width=10, fig.height=7}
ranking <- res_KD %>%
  filter(!is.na(log2FoldChange)) %>%
  arrange(desc(log2FoldChange)) %>%
  { setNames(.$log2FoldChange, .$gene) }

msig_h <- msigdbr(species = "Homo sapiens", category = "H") %>%
  dplyr::select(gs_name, gene_symbol)

gsea_res <- GSEA(
  geneList     = ranking,
  TERM2GENE    = msig_h,
  pvalueCutoff = 0.05,
  pAdjustMethod = "BH",
  verbose      = FALSE
)

write.csv(gsea_res@result,
          file.path(additional_dir, paste0("GSEA_Hallmark_siSLCLPS5_vs_siConLPS_", output_suffix, ".csv")),
          row.names = FALSE)

if (nrow(gsea_res@result) > 0) {
  dotplot(gsea_res, showCategory = 20,
          title = "GSEA Hallmark: siSLC_LPS5 vs siCon_LPS") +
    theme(axis.text.y = element_text(size = 9))
} else {
  cat("No significant GSEA results (padj < 0.05)\n")
}
```

```{r gsea_ridge, fig.width=10, fig.height=7}
if (nrow(gsea_res@result) > 0) {
  ridgeplot(gsea_res, showCategory = 15) +
    labs(title = "GSEA Ridgeplot: siSLC_LPS5 vs siCon_LPS")
}
```

# 12. ORA - GO & KEGG

```{r ora}
up_genes <- res_KD %>%
  filter(padj < 0.05, log2FoldChange > 0.58) %>%
  pull(gene)

dn_genes <- res_KD %>%
  filter(padj < 0.05, log2FoldChange < -0.58) %>%
  pull(gene)

cat("UP genes:", length(up_genes), " | DOWN genes:", length(dn_genes), "\n")
```

```{r go_bp_up, fig.width=10, fig.height=7}
if (length(up_genes) > 5) {
  go_up <- enrichGO(
    gene          = up_genes,
    OrgDb         = org.Hs.eg.db,
    keyType       = "SYMBOL",
    ont           = "BP",
    pAdjustMethod = "BH",
    pvalueCutoff  = 0.05,
    qvalueCutoff  = 0.2
  )
  write.csv(go_up@result,
            file.path(additional_dir, paste0("GO_BP_UP_siSLCLPS5_vs_siConLPS_", output_suffix, ".csv")),
            row.names = FALSE)
  if (nrow(go_up@result) > 0) {
    dotplot(go_up, showCategory = 20,
            title = "GO BP - UP in siSLC_LPS5 vs siCon_LPS")
  } else {
    cat("No significant GO BP results for UP genes\n")
  }
} else {
  cat("Not enough UP genes for ORA\n")
}
```

```{r go_bp_dn, fig.width=10, fig.height=7}
if (length(dn_genes) > 5) {
  go_dn <- enrichGO(
    gene          = dn_genes,
    OrgDb         = org.Hs.eg.db,
    keyType       = "SYMBOL",
    ont           = "BP",
    pAdjustMethod = "BH",
    pvalueCutoff  = 0.05,
    qvalueCutoff  = 0.2
  )
  write.csv(go_dn@result,
            file.path(additional_dir, paste0("GO_BP_DOWN_siSLCLPS5_vs_siConLPS_", output_suffix, ".csv")),
            row.names = FALSE)
  if (nrow(go_dn@result) > 0) {
    dotplot(go_dn, showCategory = 20,
            title = "GO BP - DOWN in siSLC_LPS5 vs siCon_LPS")
  } else {
    cat("No significant GO BP results for DOWN genes\n")
  }
} else {
  cat("Not enough DOWN genes for ORA\n")
}
```

```{r kegg_up, fig.width=10, fig.height=6}
if (length(up_genes) > 5) {
  eg_up <- bitr(up_genes, fromType = "SYMBOL",
                toType = "ENTREZID", OrgDb = org.Hs.eg.db)

  kegg_up <- enrichKEGG(
    gene          = eg_up$ENTREZID,
    organism      = "hsa",
    pAdjustMethod = "BH",
    pvalueCutoff  = 0.05
  )
  if (!is.null(kegg_up)) {
    write.csv(kegg_up@result,
              file.path(additional_dir, paste0("KEGG_UP_siSLCLPS5_vs_siConLPS_", output_suffix, ".csv")),
              row.names = FALSE)
  }
  if (!is.null(kegg_up) && nrow(kegg_up@result) > 0) {
    dotplot(kegg_up, showCategory = 20,
            title = "KEGG - UP in siSLC_LPS5 vs siCon_LPS")
  } else {
    cat("No significant KEGG results for UP genes\n")
  }
}
```

# 13. SLC Family Gene Analysis

```{r slc_extract}
slc_res <- res_KD %>%
  filter(grepl("^SLC", gene)) %>%
  arrange(padj)

cat("Total SLC family genes in results:", nrow(slc_res), "\n")
cat("SLC DEGs (padj<0.05, |log2FC|>1):",
    sum(slc_res$padj < 0.05 & abs(slc_res$log2FoldChange) > 1, na.rm = TRUE), "\n")
cat("  UP  (log2FC > 1):",
    sum(slc_res$padj < 0.05 & slc_res$log2FoldChange >  1, na.rm = TRUE), "\n")
cat("  DOWN(log2FC < -1):",
    sum(slc_res$padj < 0.05 & slc_res$log2FoldChange < -1, na.rm = TRUE), "\n\n")

slc_sig <- slc_res %>%
  filter(padj < 0.05, abs(log2FoldChange) > 1)

print(slc_sig)
write.csv(slc_res,
          file.path(additional_dir, paste0("SLC_family_DEG_siSLCLPS5_vs_siConLPS_", output_suffix, ".csv")),
          row.names = FALSE)
```

```{r slc_fisher_enrichment}
all_res <- res_KD %>% filter(!is.na(padj))
fisher_results <- list()

for (direction in c("UP", "DOWN")) {
  if (direction == "UP") {
    is_deg <- all_res$padj < 0.05 & all_res$log2FoldChange >  1
  } else {
    is_deg <- all_res$padj < 0.05 & all_res$log2FoldChange < -1
  }

  is_slc <- grepl("^SLC", all_res$gene)

  a <- sum( is_slc &  is_deg, na.rm = TRUE)
  b <- sum( is_slc & !is_deg, na.rm = TRUE)
  c <- sum(!is_slc &  is_deg, na.rm = TRUE)
  d <- sum(!is_slc & !is_deg, na.rm = TRUE)

  ct <- matrix(c(a, c, b, d), nrow = 2,
               dimnames = list(c("SLC", "non-SLC"), c("DEG", "non-DEG")))

  ft <- fisher.test(ct)

  fisher_results[[direction]] <- data.frame(
    Direction    = direction,
    SLC_DEG      = a,
    SLC_nonDEG   = b,
    nonSLC_DEG   = c,
    nonSLC_nonDEG= d,
    OddsRatio    = round(ft$estimate, 4),
    p_value      = ft$p.value,
    CI_lower     = round(ft$conf.int[1], 4),
    CI_upper     = round(ft$conf.int[2], 4),
    Significant  = ft$p.value < 0.05
  )

  cat(sprintf("[SLC enrichment - %s DEGs]\n", direction))
  print(ct)
  cat(sprintf("  Odds Ratio : %.4f (95%% CI: %.4f - %.4f)\n",
              ft$estimate, ft$conf.int[1], ft$conf.int[2]))
  cat(sprintf("  p-value    : %.4e\n\n", ft$p.value))
}

enrich_df <- do.call(rbind, fisher_results)
rownames(enrich_df) <- NULL
print(enrich_df)
write.csv(enrich_df,
          file.path(additional_dir, paste0("SLC_fisher_enrichment_", output_suffix, ".csv")),
          row.names = FALSE)
```

```{r slc_per_gene_fisher}
ctrl_cols <- grep("Cont_L", colnames(counts_filt), value = TRUE)
kd_cols   <- grep("SLC_L",  colnames(counts_filt), value = TRUE)

cat("siCon_LPS samples :", paste(ctrl_cols, collapse = ", "), "\n")
cat("siSLC_LPS5 samples:", paste(kd_cols,   collapse = ", "), "\n\n")

lib_ctrl <- sum(counts_filt[, ctrl_cols])
lib_kd   <- sum(counts_filt[, kd_cols])

slc_genes_in_counts <- rownames(counts_filt)[grepl("^SLC", rownames(counts_filt))]
slc_counts <- counts_filt[slc_genes_in_counts, c(ctrl_cols, kd_cols), drop = FALSE]

per_gene_fisher <- lapply(slc_genes_in_counts, function(gene) {
  g_ctrl <- sum(slc_counts[gene, ctrl_cols])
  g_kd   <- sum(slc_counts[gene, kd_cols])

  ct <- matrix(
    c(g_ctrl, g_kd,
      lib_ctrl - g_ctrl, lib_kd - g_kd),
    nrow = 2,
    dimnames = list(c("siCon_LPS", "siSLC_LPS5"),
                    c("this_gene", "other_genes"))
  )

  ft <- tryCatch(fisher.test(ct), error = function(e) NULL)
  if (is.null(ft)) return(NULL)

  data.frame(
    gene         = gene,
    count_ctrl   = g_ctrl,
    count_kd     = g_kd,
    OddsRatio    = round(ft$estimate, 4),
    p_value      = ft$p.value,
    stringsAsFactors = FALSE
  )
})

per_gene_df <- do.call(rbind, Filter(Negate(is.null), per_gene_fisher))
per_gene_df$p_adj <- p.adjust(per_gene_df$p_value, method = "BH")
per_gene_df$log2OR <- log2(per_gene_df$OddsRatio + 1e-6)

per_gene_df <- per_gene_df %>%
  left_join(slc_res %>% select(gene, log2FoldChange, padj), by = "gene") %>%
  arrange(p_adj)

cat("SLC genes significant by Fisher (p_adj < 0.05):",
    sum(per_gene_df$p_adj < 0.05, na.rm = TRUE), "\n\n")

print(head(per_gene_df, 20))
write.csv(per_gene_df,
          file.path(additional_dir, paste0("SLC_per_gene_fisher_siSLCLPS5_vs_siConLPS_", output_suffix, ".csv")),
          row.names = FALSE)
```

```{r slc_volcano, fig.width=10, fig.height=7}
plot_df <- res_KD %>%
  mutate(
    is_slc = grepl("^SLC", gene),
    sig    = !is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 0.58,
    label  = ifelse(is_slc & sig, gene, NA)
  )

ggplot(plot_df, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(data = filter(plot_df, !is_slc),
             color = "grey80", size = 0.8, alpha = 0.5) +
  geom_point(data = filter(plot_df,  is_slc & !sig),
             color = "#90CAF9", size = 1.5, alpha = 0.8) +
  geom_point(data = filter(plot_df,  is_slc &  sig),
             aes(color = log2FoldChange > 0), size = 2.5) +
  scale_color_manual(values = c("TRUE" = "#D32F2F", "FALSE" = "#1565C0"),
                     labels = c("TRUE" = "SLC UP", "FALSE" = "SLC DOWN")) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = c(-0.58, 0.58), linetype = "dashed", color = "grey40") +
  geom_text_repel(aes(label = label), size = 3, max.overlaps = 20,
                  segment.color = "grey50") +
  labs(
    title    = "Volcano Plot: siSLC_LPS5 vs siCon_LPS",
    subtitle = "SLC family genes highlighted",
    x        = "log2 Fold Change",
    y        = "-log10(padj)",
    color    = ""
  ) +
  theme_bw() +
  theme(legend.position = "top")
```

```{r slc_bar, fig.width=10, fig.height=6}
slc_sig_bar <- per_gene_df %>%
  filter(p_adj < 0.05) %>%
  arrange(log2FoldChange)

if (nrow(slc_sig_bar) > 0) {
  slc_sig_bar$gene <- factor(slc_sig_bar$gene, levels = slc_sig_bar$gene)

  p <- ggplot(slc_sig_bar, aes(x = gene, y = log2FoldChange,
                               fill = log2FoldChange > 0)) +
    geom_bar(stat = "identity") +
    geom_text(aes(label = sprintf("p=%.3f", p_adj),
                  y = ifelse(log2FoldChange > 0,
                             log2FoldChange + 0.05,
                             log2FoldChange - 0.05)),
              hjust = ifelse(slc_sig_bar$log2FoldChange > 0, 0, 1),
              size = 3) +
    coord_flip() +
    scale_fill_manual(values = c("TRUE" = "#D32F2F", "FALSE" = "#1565C0"),
                      labels = c("TRUE" = "UP", "FALSE" = "DOWN")) +
    labs(
      title = "Significant SLC Family Genes (Fisher p_adj < 0.05)",
      subtitle = "siSLC_LPS5 vs siCon_LPS",
      x = "", y = "log2 Fold Change (DESeq2)", fill = ""
    ) +
    theme_bw() +
    theme(legend.position = "top")

  print(p)
  ggsave(file.path(additional_dir, paste0("slc_sig_barplot_", output_suffix, ".png")),
         plot = p, width = 12, height = 8, dpi = 300)
} else {
  cat("No SLC genes significant after BH correction (p_adj < 0.05)\n")

  top20_slc <- per_gene_df %>% slice_head(n = 20)
  top20_slc$gene <- factor(
    top20_slc$gene,
    levels = top20_slc$gene[order(top20_slc$log2FoldChange)]
  )

  p <- ggplot(top20_slc, aes(x = gene, y = log2FoldChange,
                             fill = log2FoldChange > 0)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    scale_fill_manual(values = c("TRUE" = "#D32F2F", "FALSE" = "#1565C0"),
                      labels = c("TRUE" = "UP", "FALSE" = "DOWN")) +
    labs(title = "Top 20 SLC Genes by Fisher p-value (uncorrected)",
         x = "", y = "log2 Fold Change", fill = "") +
    theme_bw()

  print(p)
  ggsave(file.path(additional_dir, paste0("slc_top20_barplot_", output_suffix, ".png")),
         plot = p, width = 12, height = 8, dpi = 300)
}
```

```{r slc_heatmap, fig.width=10, fig.height=8}
slc_in_vst <- rownames(vst_mat)[grepl("^SLC", rownames(vst_mat))]

slc_show <- per_gene_df %>%
  { if (sum(.$p_adj < 0.05) >= 5)
      filter(., p_adj < 0.05)
    else
      slice_head(., n = 30) } %>%
  pull(gene) %>%
  intersect(slc_in_vst)

if (length(slc_show) >= 2) {
  vst_slc <- vst_mat[slc_show, c(ctrl_cols, kd_cols), drop = FALSE]

  ann_col_slc <- data.frame(
    group = c(rep("siCon_LPS", length(ctrl_cols)),
              rep("siSLC_LPS5", length(kd_cols)))
  )
  rownames(ann_col_slc) <- c(ctrl_cols, kd_cols)

  group_colors_slc <- list(group = c(
    siCon_LPS  = "#D6604D",
    siSLC_LPS5 = "#74C476"
  ))

  pheatmap(
    vst_slc,
    scale             = "row",
    annotation_col    = ann_col_slc,
    annotation_colors = group_colors_slc,
    clustering_method = "ward.D2",
    color             = colorRampPalette(c("#2166AC", "white", "#B2182B"))(200),
    show_rownames     = TRUE,
    fontsize_row      = 8,
    main              = "SLC Family Gene Expression Heatmap\n(siCon_LPS vs siSLC_LPS5)"
  )

  pdf(file.path(additional_dir, paste0("SLC_heatmap_", output_suffix, ".pdf")),
      width = 10, height = 20)
  pheatmap(
    vst_slc,
    scale             = "row",
    annotation_col    = ann_col_slc,
    annotation_colors = group_colors_slc,
    clustering_method = "ward.D2",
    color             = colorRampPalette(c("#2166AC", "white", "#B2182B"))(200),
    show_rownames     = TRUE,
    fontsize_row      = 8,
    main              = "SLC Family Gene Expression Heatmap\n(siCon_LPS vs siSLC_LPS5)"
  )
  dev.off()
} else {
  cat("Not enough SLC genes to draw heatmap\n")
}
```

# 14. LPS-induced Genes Reversed by siSLC7A5

```{r lps_up_kd_down_overlap, fig.width=9, fig.height=5}
lps_up_genes <- res_LPS %>%
  filter(!is.na(padj), padj < 0.05, log2FoldChange > 0.58) %>%
  pull(gene) %>%
  unique()

kd_down_genes <- res_KD %>%
  filter(!is.na(padj), padj < 0.05, log2FoldChange < -0.58) %>%
  pull(gene) %>%
  unique()

overlap_genes <- intersect(lps_up_genes, kd_down_genes)
union_genes <- union(lps_up_genes, kd_down_genes)
all_tested_genes <- intersect(
  res_LPS %>% filter(!is.na(padj)) %>% pull(gene),
  res_KD %>% filter(!is.na(padj)) %>% pull(gene)
)

overlap_table <- matrix(
  c(
    length(intersect(lps_up_genes, kd_down_genes)),
    length(setdiff(lps_up_genes, kd_down_genes)),
    length(setdiff(kd_down_genes, lps_up_genes)),
    length(setdiff(all_tested_genes, union_genes))
  ),
  nrow = 2,
  byrow = TRUE,
  dimnames = list(
    LPS_up = c("yes", "no"),
    KD_down = c("yes", "no")
  )
)

overlap_fisher <- fisher.test(overlap_table, alternative = "greater")

overlap_summary <- tibble(
  category = c(
    "siCon_LPS_up_vs_siCon",
    "siSLC_LPS5_down_vs_siCon_LPS",
    "overlap",
    "union"
  ),
  n_genes = c(
    length(lps_up_genes),
    length(kd_down_genes),
    length(overlap_genes),
    length(union_genes)
  )
) %>%
  mutate(
    percent_of_LPS_up = ifelse(length(lps_up_genes) > 0,
                               100 * n_genes / length(lps_up_genes), NA_real_),
    percent_of_KD_down = ifelse(length(kd_down_genes) > 0,
                                100 * n_genes / length(kd_down_genes), NA_real_)
  )

print(overlap_summary)
print(overlap_table)
cat("Fisher exact test for overlap enrichment:\n")
cat("  odds ratio:", overlap_fisher$estimate, "\n")
cat("  p-value   :", overlap_fisher$p.value, "\n")

overlap_enrichment <- tibble(
  universe_genes = length(all_tested_genes),
  lps_up_genes = length(lps_up_genes),
  kd_down_genes = length(kd_down_genes),
  overlap_genes = length(overlap_genes),
  odds_ratio = unname(overlap_fisher$estimate),
  fisher_p_value = overlap_fisher$p.value
)

overlap_membership <- tibble(gene = union_genes) %>%
  mutate(
    siCon_LPS_up_vs_siCon = gene %in% lps_up_genes,
    siSLC_LPS5_down_vs_siCon_LPS = gene %in% kd_down_genes,
    overlap = siCon_LPS_up_vs_siCon & siSLC_LPS5_down_vs_siCon_LPS
  ) %>%
  left_join(
    res_LPS %>%
      select(gene, LPS_log2FC = log2FoldChange, LPS_padj = padj),
    by = "gene"
  ) %>%
  left_join(
    res_KD %>%
      select(gene, KD_log2FC = log2FoldChange, KD_padj = padj),
    by = "gene"
  ) %>%
  arrange(desc(overlap), desc(LPS_log2FC), KD_log2FC)

write.csv(overlap_summary,
          file.path(additional_dir, paste0("LPS_up_KD_down_overlap_summary_", output_suffix, ".csv")),
          row.names = FALSE)
write.csv(overlap_enrichment,
          file.path(additional_dir, paste0("LPS_up_KD_down_overlap_enrichment_", output_suffix, ".csv")),
          row.names = FALSE)
write.csv(overlap_membership,
          file.path(additional_dir, paste0("LPS_up_KD_down_overlap_genes_", output_suffix, ".csv")),
          row.names = FALSE)

plot_counts <- tibble(
  intersection = c("LPS up only", "Overlap", "KD down only"),
  n_genes = c(
    length(setdiff(lps_up_genes, kd_down_genes)),
    length(overlap_genes),
    length(setdiff(kd_down_genes, lps_up_genes))
  )
) %>%
  mutate(intersection = factor(intersection,
                               levels = c("LPS up only", "Overlap", "KD down only")))

ggplot(plot_counts, aes(x = intersection, y = n_genes, fill = intersection)) +
  geom_col(width = 0.7, color = "grey25") +
  geom_text(aes(label = n_genes), vjust = -0.4, size = 4) +
  scale_fill_manual(values = c(
    "LPS up only" = "#D95F02",
    "Overlap" = "#1B9E77",
    "KD down only" = "#7570B3"
  )) +
  labs(
    title = "Overlap of LPS-induced and siSLC7A5-reduced genes",
    subtitle = paste0(
      "Overlap = ", length(overlap_genes), " genes; ",
      ifelse(length(lps_up_genes) > 0,
             sprintf("%.1f", 100 * length(overlap_genes) / length(lps_up_genes)),
             "NA"),
      "% of siCon_LPS-up genes; Fisher p = ",
      formatC(overlap_fisher$p.value, format = "e", digits = 2)
    ),
    x = "",
    y = "Number of genes",
    fill = ""
  ) +
  theme_bw() +
  theme(legend.position = "none")
```

```{r lps_up_kd_down_upset_like, fig.width=8, fig.height=5}
upset_like_df <- tibble(
  intersection = c("LPS up only", "Overlap", "KD down only"),
  n_genes = c(
    length(setdiff(lps_up_genes, kd_down_genes)),
    length(overlap_genes),
    length(setdiff(kd_down_genes, lps_up_genes))
  ),
  LPS_up = c(TRUE, TRUE, FALSE),
  KD_down = c(FALSE, TRUE, TRUE)
) %>%
  arrange(desc(n_genes), intersection) %>%
  mutate(intersection = factor(intersection, levels = intersection))

bar_df <- upset_like_df %>%
  select(intersection, n_genes)

plot_max <- max(bar_df$n_genes, 1)
label_x <- levels(upset_like_df$intersection)[1]

dot_df <- upset_like_df %>%
  tidyr::pivot_longer(c(LPS_up, KD_down),
                      names_to = "gene_set",
                      values_to = "included") %>%
  mutate(
    gene_set = factor(gene_set,
                      levels = c("KD_down", "LPS_up"),
                      labels = c("siSLC_LPS5 down", "siCon_LPS up"))
  )

ggplot() +
  geom_col(data = bar_df,
           aes(x = intersection, y = n_genes),
           width = 0.65, fill = "#4D4D4D") +
  geom_text(data = bar_df,
            aes(x = intersection, y = n_genes, label = n_genes),
            vjust = -0.4, size = 4) +
  geom_point(data = dot_df,
             aes(x = intersection,
                 y = -0.16 * plot_max -
                   as.numeric(gene_set) * 0.12 * plot_max,
                 alpha = included),
             size = 4, color = "#1B9E77") +
  geom_line(data = filter(dot_df, included),
            aes(x = intersection,
                y = -0.16 * plot_max -
                  as.numeric(gene_set) * 0.12 * plot_max,
                group = intersection),
            linewidth = 0.8, color = "#1B9E77") +
  scale_alpha_manual(values = c("TRUE" = 1, "FALSE" = 0.15), guide = "none") +
  scale_y_continuous(
    limits = c(-0.48 * plot_max, plot_max * 1.15),
    breaks = scales::pretty_breaks(),
    labels = function(x) ifelse(x < 0, "", x),
    expand = expansion(mult = c(0, 0.02))
  ) +
  annotate("text",
           x = label_x,
           y = -0.16 * plot_max - c(1, 2) * 0.12 * plot_max,
           label = c("siSLC_LPS5 down", "siCon_LPS up"),
           hjust = 1.25, size = 3.5) +
  labs(
    title = "UpSet-like overlap: LPS-up genes reversed by siSLC7A5 KD",
    subtitle = "Thresholds: padj < 0.05, LPS log2FC > 0.58, KD log2FC < -0.58",
    x = "",
    y = "Intersection size"
  ) +
  theme_bw() +
  theme(
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(angle = 20, hjust = 1)
  )
```

```{r lps_up_kd_down_venn, fig.width=7, fig.height=5}
venn_counts <- tibble(
  region = c("LPS up only", "Overlap", "KD down only"),
  n_genes = c(
    length(setdiff(lps_up_genes, kd_down_genes)),
    length(overlap_genes),
    length(setdiff(kd_down_genes, lps_up_genes))
  )
)

circle_df <- bind_rows(
  tibble(
    set = "siCon_LPS up",
    angle = seq(0, 2 * pi, length.out = 300),
    center_x = -0.65,
    center_y = 0,
    radius = 1
  ),
  tibble(
    set = "siSLC_LPS5 down",
    angle = seq(0, 2 * pi, length.out = 300),
    center_x = 0.65,
    center_y = 0,
    radius = 1
  )
) %>%
  mutate(
    x = center_x + radius * cos(angle),
    y = center_y + radius * sin(angle)
  )

label_df <- tibble(
  x = c(-1.05, 0, 1.05),
  y = c(0, 0, 0),
  label = c(
    paste0("LPS up only\n", venn_counts$n_genes[venn_counts$region == "LPS up only"]),
    paste0("Overlap\n", venn_counts$n_genes[venn_counts$region == "Overlap"]),
    paste0("KD down only\n", venn_counts$n_genes[venn_counts$region == "KD down only"])
  )
)

set_label_df <- tibble(
  x = c(-0.95, 0.95),
  y = c(1.15, 1.15),
  label = c(
    paste0("siCon_LPS up\nn = ", length(lps_up_genes)),
    paste0("siSLC_LPS5 down\nn = ", length(kd_down_genes))
  )
)

ggplot(circle_df, aes(x = x, y = y, fill = set, color = set, group = set)) +
  geom_polygon(alpha = 0.28, linewidth = 1.1) +
  geom_text(data = label_df,
            aes(x = x, y = y, label = label),
            inherit.aes = FALSE, size = 4.2, fontface = "bold") +
  geom_text(data = set_label_df,
            aes(x = x, y = y, label = label),
            inherit.aes = FALSE, size = 3.8) +
  scale_fill_manual(values = c("siCon_LPS up" = "#D95F02",
                               "siSLC_LPS5 down" = "#7570B3")) +
  scale_color_manual(values = c("siCon_LPS up" = "#D95F02",
                                "siSLC_LPS5 down" = "#7570B3")) +
  coord_equal(xlim = c(-2, 2), ylim = c(-1.25, 1.45), expand = FALSE) +
  labs(
    title = "Venn diagram: LPS-up genes reversed by siSLC7A5 KD",
    subtitle = paste0(
      "Overlap = ", length(overlap_genes), " genes; Fisher p = ",
      formatC(overlap_fisher$p.value, format = "e", digits = 2)
    ),
    x = NULL,
    y = NULL
  ) +
  theme_void() +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )
```

# 15. Transcript-level DESeq2 Analysis

```{r transcript_deseq2}
tx_counts_mat <- tx_expr$counts
tx_sample_info <- make_sample_info(colnames(tx_counts_mat))

tx_de <- run_deseq2(tx_counts_mat, tx_sample_info)
tx_dds <- tx_de$dds
tx_counts_filt <- tx_de$counts_filt

cat("Transcripts after pre-filtering:", nrow(tx_counts_filt), "\n")
cat("Transcript DESeq2 design:", as.character(design(tx_dds)), "\n")

tx_res_LPS <- get_results(tx_dds, contrast = c("group", "siCon_LPS", "siCon"))
tx_res_KD <- get_results(tx_dds, contrast = c("group", "siSLC_LPS5", "siCon_LPS"))
tx_res_overall <- get_results(tx_dds, contrast = c("group", "siSLC_LPS5", "siCon"))

summarize_deg(tx_res_LPS, "Transcript: siCon_LPS vs siCon")
summarize_deg(tx_res_KD, "Transcript: siSLC_LPS5 vs siCon_LPS")
summarize_deg(tx_res_overall, "Transcript: siSLC_LPS5 vs siCon")

write.csv(tx_res_LPS,
          file.path(additional_dir, paste0("Transcript_DEG_siConLPS_vs_siCon_", output_suffix, ".csv")),
          row.names = FALSE)
write.csv(tx_res_KD,
          file.path(additional_dir, paste0("Transcript_DEG_siSLCLPS5_vs_siConLPS_", output_suffix, ".csv")),
          row.names = FALSE)
write.csv(tx_res_overall,
          file.path(additional_dir, paste0("Transcript_DEG_siSLCLPS5_vs_siCon_", output_suffix, ".csv")),
          row.names = FALSE)
```

```{r transcript_pca, fig.width=8, fig.height=6}
tx_vst <- vst(tx_dds, blind = TRUE)
tx_pca_data <- plotPCA(tx_vst, intgroup = "group", returnData = TRUE)
tx_pct_var  <- round(100 * attr(tx_pca_data, "percentVar"))

ggplot(tx_pca_data, aes(x = PC1, y = PC2, color = group, label = name)) +
  geom_point(size = 4) +
  geom_text_repel(size = 3) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "PCA - VST normalized transcript counts",
    x = paste0("PC1: ", tx_pct_var[1], "% variance"),
    y = paste0("PC2: ", tx_pct_var[2], "% variance")
  ) +
  theme_bw()
```

```{r transcript_heatmap, fig.width=10, fig.height=8}
tx_vst_mat <- assay(tx_vst)
tx_sig <- unique(c(
  tx_res_LPS     %>% filter(padj < 0.05, abs(log2FoldChange) > 0.58) %>% pull(gene),
  tx_res_KD      %>% filter(padj < 0.05, abs(log2FoldChange) > 0.58) %>% pull(gene),
  tx_res_overall %>% filter(padj < 0.05, abs(log2FoldChange) > 0.58) %>% pull(gene)
))

tx_vst_sig <- tx_vst_mat[rownames(tx_vst_mat) %in% tx_sig, , drop = FALSE]

if (nrow(tx_vst_sig) > 100) {
  tx_var <- apply(tx_vst_sig, 1, var)
  tx_vst_sig <- tx_vst_sig[order(tx_var, decreasing = TRUE)[1:100], , drop = FALSE]
}

if (nrow(tx_vst_sig) >= 2) {
  pheatmap(
    tx_vst_sig,
    scale              = "row",
    annotation_col     = ann_col,
    annotation_colors  = group_colors,
    show_rownames      = nrow(tx_vst_sig) <= 50,
    clustering_method  = "ward.D2",
    color              = colorRampPalette(c("#2166AC", "white", "#B2182B"))(200),
    main               = paste0("Top transcript DEG Heatmap (n=", nrow(tx_vst_sig), ")")
  )
} else {
  cat("Not enough significant transcripts for transcript heatmap\n")
}
```

# 16. Session Info

```{r session_info}
sessionInfo()
```
