---
title: "Supplementary Figure S3 - Shared NF-kB Target Genes"
author: "SLC7A5 monocyte project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r repository-setup, include=FALSE}
.slc_start <- Sys.getenv("SLC7A5_REPO_ROOT", unset = getwd())
.slc_repo <- normalizePath(.slc_start, mustWork = TRUE)
while (!file.exists(file.path(.slc_repo, "R", "paths.R"))) {
  .slc_parent <- dirname(.slc_repo)
  if (identical(.slc_parent, .slc_repo)) stop("Start inside the repository or set SLC7A5_REPO_ROOT.")
  .slc_repo <- .slc_parent
}
source(file.path(.slc_repo, "R", "paths.R"))
slc_setup("nfkb_shared")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

# 1. Analysis Goal

This analysis defines shared NF-kB target genes that are:

1. Upregulated by LPS in the bulk RNA-seq comparison.
2. Upregulated in the MASLD SLC7A5-high myeloid/monocyte signature.
3. Annotated as NF-kB pathway target genes.

The resulting gene set is then tested in the bulk RNA-seq samples:

- siControl
- siControl + LPS
- siSLC7A5 + LPS

Expected pattern:

- siControl: low
- siControl + LPS: high
- siSLC7A5 + LPS: reduced

# 2. Libraries and Parameters

```{r libraries}
required_pkgs <- c(
  "dplyr", "tidyr", "readr", "ggplot2", "pheatmap", "tibble",
  "stringr", "purrr", "scales"
)

missing_pkgs <- required_pkgs[!vapply(required_pkgs, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_pkgs) > 0) {
  stop(
    "Missing required R packages: ",
    paste(missing_pkgs, collapse = ", "),
    "\nInstall them before knitting this Rmd."
  )
}

library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(pheatmap)
library(tibble)
library(stringr)
library(purrr)
library(scales)
```

```{r parameters}
project_dir <- slc_output_dir("02_SLC7A5_mono_Journal")
bulk_dir <- slc_output_dir("02_SLC7A5_mono_Journal/04_bulkRNAseq")
out_dir <- slc_output_dir("02_SLC7A5_mono_Journal/04_bulkRNAseq/SupplFigS3_shared_NFKB_targets")
lps_deg_file <- slc_input("bulk_nosi_lps_deg")
masld_slc7a5_high_deg_file <- slc_input("scrna_myeloid_deg")
tpm_file <- slc_input("bulk_tpm")
nfkb_local_gsea_file <- slc_input("scrna_myeloid_gsea", required = FALSE)

# DEG thresholds.
# For LPS bulk RNA-seq, keep a stricter fold-change threshold by default.
lps_padj_cutoff <- 0.05
lps_log2fc_min <- 1

# For scRNA-seq SLC7A5-high signature, logfoldchanges may be natural-log-like
# depending on the upstream tool. Keep this less strict by default.
masld_padj_cutoff <- 0.05
masld_logfc_min <- 0.25

# Heatmap/sample options.
exclude_samples <- c("siCon_1")
zscore_cap <- 2

# Candidate genes for S3B. The final plot keeps genes present in the shared
# gene set and in the bulk expression matrix, then prioritizes genes showing
# LPS-up and siSLC7A5-LPS-down behavior.
representative_candidates <- c(
  "IL1B", "TNF", "CXCL8", "CCL2", "NFKBIA", "TNFAIP3", "ICAM1", "PTGS2",
  "CXCL1", "CXCL2", "CXCL3", "IL6", "BIRC3", "SOD2", "JUNB", "FOS",
  "DUSP1", "SOCS3", "TNFAIP6", "PTX3"
)

stopifnot(file.exists(lps_deg_file))
stopifnot(file.exists(masld_slc7a5_high_deg_file))
stopifnot(file.exists(tpm_file))
```

# 3. Helper Functions

```{r helper-functions}
standardize_deg <- function(df) {
  nm <- names(df)

  gene_col <- dplyr::case_when(
    "gene" %in% nm ~ "gene",
    "gene_name" %in% nm ~ "gene_name",
    "Gene" %in% nm ~ "Gene",
    "names" %in% nm ~ "names",
    TRUE ~ NA_character_
  )

  lfc_col <- dplyr::case_when(
    "log2FoldChange" %in% nm ~ "log2FoldChange",
    "logfoldchanges" %in% nm ~ "logfoldchanges",
    "avg_log2FC" %in% nm ~ "avg_log2FC",
    "avg_logFC" %in% nm ~ "avg_logFC",
    "logFC" %in% nm ~ "logFC",
    TRUE ~ NA_character_
  )

  padj_col <- dplyr::case_when(
    "padj" %in% nm ~ "padj",
    "pvals_adj" %in% nm ~ "pvals_adj",
    "p_val_adj" %in% nm ~ "p_val_adj",
    "FDR" %in% nm ~ "FDR",
    TRUE ~ NA_character_
  )

  if (is.na(gene_col) || is.na(lfc_col) || is.na(padj_col)) {
    stop(
      "Cannot identify gene/logFC/adjusted-p columns. Columns are: ",
      paste(nm, collapse = ", ")
    )
  }

  df %>%
    transmute(
      gene = toupper(as.character(.data[[gene_col]])),
      logFC = suppressWarnings(as.numeric(.data[[lfc_col]])),
      padj = suppressWarnings(as.numeric(.data[[padj_col]]))
    ) %>%
    filter(!is.na(gene), gene != "", !is.na(logFC), !is.na(padj)) %>%
    distinct(gene, .keep_all = TRUE)
}

row_zscore <- function(mat) {
  z <- t(scale(t(mat)))
  z[is.na(z)] <- 0
  z
}

mean_sem <- function(x) {
  x <- x[!is.na(x)]
  tibble(
    mean = mean(x),
    sem = stats::sd(x) / sqrt(length(x))
  )
}

sample_to_group <- function(sample) {
  dplyr::case_when(
    grepl("^siCon_LPS", sample) ~ "siControl + LPS",
    grepl("^siSLC_LPS5", sample) ~ "siSLC7A5 + LPS",
    grepl("^siCon", sample) ~ "siControl",
    TRUE ~ sample
  )
}

safe_t_test <- function(df, value_col, group_a, group_b) {
  x <- df %>% filter(group == group_a) %>% pull(all_of(value_col))
  y <- df %>% filter(group == group_b) %>% pull(all_of(value_col))
  if (length(x) < 2 || length(y) < 2) return(NA_real_)
  stats::t.test(x, y)$p.value
}

p_to_star <- function(p) {
  dplyr::case_when(
    is.na(p) ~ "n.s.",
    p < 0.001 ~ "***",
    p < 0.01 ~ "**",
    p < 0.05 ~ "*",
    TRUE ~ "n.s."
  )
}
```

# 4. Define NF-kB Target Genes

The preferred source is MSigDB Hallmark `HALLMARK_TNFA_SIGNALING_VIA_NFKB` through `msigdbr`, if installed. If `msigdbr` is unavailable, this Rmd falls back to local GSEA leading-edge genes and a small curated NF-kB target list.

```{r nfkb-target-genes}
manual_nfkb_targets <- toupper(c(
  "IL1B", "TNF", "CXCL8", "CCL2", "NFKBIA", "TNFAIP3", "ICAM1", "PTGS2",
  "IL6", "CXCL1", "CXCL2", "CXCL3", "CXCL10", "CCL3", "CCL4", "CCL5",
  "NFKB1", "NFKB2", "RELA", "REL", "RELB", "NFKBIB", "NFKBIE",
  "BIRC2", "BIRC3", "TRAF1", "TRAF2", "TRAF3", "MAP3K8", "MYD88",
  "TLR2", "TLR4", "CD14", "LY96", "IRAK1", "IRAK2", "TANK",
  "SOCS3", "DUSP1", "DUSP2", "JUN", "JUNB", "FOS", "FOSB", "ATF3",
  "EGR1", "EGR2", "EGR3", "IER2", "IER3", "ZFP36", "KLF6", "KLF9",
  "NFKBIZ", "TNFAIP2", "TNFAIP6", "TNIP1", "TNIP2", "ZC3H12A",
  "SOD2", "NAMPT", "PLAUR", "PTX3", "SERPINE1", "GADD45B", "G0S2",
  "MCL1", "BCL2A1", "CD44", "CD69", "CEBPB", "CEBPD", "HBEGF",
  "AREG", "VEGFA", "LIF", "IL18", "IFNGR2", "KYNU", "BCL6", "BTG1",
  "BTG2", "RHOB", "NR4A1", "NR4A2", "MARCKS", "SAT1", "SLC2A3"
))

msig_nfkb_targets <- character(0)
if (requireNamespace("msigdbr", quietly = TRUE)) {
  msig_nfkb_targets <- msigdbr::msigdbr(
    species = "Homo sapiens",
    category = "H"
  ) %>%
    filter(gs_name == "HALLMARK_TNFA_SIGNALING_VIA_NFKB") %>%
    pull(gene_symbol) %>%
    toupper() %>%
    unique()
}

local_nfkb_targets <- character(0)
if (file.exists(nfkb_local_gsea_file)) {
  local_gsea <- suppressMessages(readr::read_csv(nfkb_local_gsea_file))
  local_nfkb_targets <- local_gsea %>%
    filter(str_detect(tolower(Term), "nf-kb|nfkb|tnf-alpha")) %>%
    pull(Lead_genes) %>%
    str_split(";") %>%
    unlist() %>%
    toupper() %>%
    unique()
}

nfkb_targets <- unique(c(msig_nfkb_targets, local_nfkb_targets, manual_nfkb_targets))

nfkb_source_summary <- tibble(
  source = c("MSigDB_Hallmark_msigdbr", "local_GSEA_leading_edge", "manual_curated_fallback", "union"),
  n_genes = c(
    length(msig_nfkb_targets),
    length(local_nfkb_targets),
    length(manual_nfkb_targets),
    length(nfkb_targets)
  )
)

readr::write_csv(nfkb_source_summary, file.path(out_dir, "S3_NFKB_target_source_summary.csv"))
readr::write_csv(tibble(gene = sort(nfkb_targets)), file.path(out_dir, "S3_NFKB_target_genes_union.csv"))

nfkb_source_summary
```

# 5. Create Shared NF-kB Target Gene List

```{r shared-gene-list}
lps_deg <- suppressMessages(readr::read_csv(lps_deg_file)) %>%
  standardize_deg()

masld_deg <- suppressMessages(readr::read_csv(masld_slc7a5_high_deg_file)) %>%
  standardize_deg()

lps_up_genes <- lps_deg %>%
  filter(padj < lps_padj_cutoff, logFC > lps_log2fc_min) %>%
  pull(gene) %>%
  unique()

masld_slc7a5_high_up_genes <- masld_deg %>%
  filter(padj < masld_padj_cutoff, logFC > masld_logfc_min) %>%
  pull(gene) %>%
  unique()

shared_nfkb_genes <- Reduce(
  intersect,
  list(lps_up_genes, masld_slc7a5_high_up_genes, nfkb_targets)
) %>%
  sort()

shared_gene_table <- tibble(gene = shared_nfkb_genes) %>%
  left_join(lps_deg %>% rename(lps_logFC = logFC, lps_padj = padj), by = "gene") %>%
  left_join(masld_deg %>% rename(masld_logFC = logFC, masld_padj = padj), by = "gene") %>%
  arrange(desc(lps_logFC), masld_padj)

gene_set_summary <- tibble(
  category = c(
    "LPS_up_bulk",
    "MASLD_SLC7A5_high_myeloid_up",
    "NF-kB_targets",
    "Shared_NF-kB_targets"
  ),
  n_genes = c(
    length(lps_up_genes),
    length(masld_slc7a5_high_up_genes),
    length(nfkb_targets),
    length(shared_nfkb_genes)
  )
)

readr::write_csv(gene_set_summary, file.path(out_dir, "S3_shared_NFKB_gene_set_summary.csv"))
readr::write_csv(shared_gene_table, file.path(out_dir, "S3_shared_NFKB_target_genes.csv"))

gene_set_summary
```

```{r show-shared-genes}
if (length(shared_nfkb_genes) < 3) {
  warning(
    "Shared NF-kB target gene set has fewer than 3 genes. ",
    "Consider relaxing lps_log2fc_min or masld_logfc_min in the parameter block."
  )
}

shared_gene_table
```

# 6. Load Bulk Expression Matrix

```{r bulk-expression}
tpm_raw <- suppressMessages(readr::read_tsv(tpm_file))

gene_col <- dplyr::case_when(
  "gene_name" %in% names(tpm_raw) ~ "gene_name",
  "gene" %in% names(tpm_raw) ~ "gene",
  "gene_id" %in% names(tpm_raw) ~ "gene_id",
  TRUE ~ names(tpm_raw)[1]
)

expr_tbl <- tpm_raw %>%
  mutate(gene = toupper(as.character(.data[[gene_col]]))) %>%
  select(gene, where(is.numeric)) %>%
  group_by(gene) %>%
  summarise(across(everything(), ~ mean(.x, na.rm = TRUE)), .groups = "drop") %>%
  filter(!is.na(gene), gene != "")

sample_cols <- setdiff(names(expr_tbl), "gene")
sample_cols <- setdiff(sample_cols, exclude_samples)

expr_mat <- expr_tbl %>%
  select(gene, all_of(sample_cols)) %>%
  column_to_rownames("gene") %>%
  as.matrix()

log_expr_mat <- log2(expr_mat + 1)

sample_info <- tibble(
  sample = colnames(log_expr_mat),
  group = sample_to_group(sample)
) %>%
  mutate(group = factor(group, levels = c("siControl", "siControl + LPS", "siSLC7A5 + LPS")))

sample_info
```

# 7. S3A - Heatmap of Shared NF-kB Targets

```{r heatmap-data}
heatmap_genes <- intersect(shared_nfkb_genes, rownames(log_expr_mat))

missing_heatmap_genes <- setdiff(shared_nfkb_genes, rownames(log_expr_mat))
if (length(missing_heatmap_genes) > 0) {
  readr::write_csv(
    tibble(gene = missing_heatmap_genes),
    file.path(out_dir, "S3_shared_NFKB_genes_missing_from_bulk_TPM.csv")
  )
}

if (length(heatmap_genes) < 2) {
  stop("Fewer than 2 shared NF-kB target genes are present in the bulk TPM matrix.")
}

heatmap_mat <- log_expr_mat[heatmap_genes, , drop = FALSE]
heatmap_z <- row_zscore(heatmap_mat)
heatmap_z <- pmax(pmin(heatmap_z, zscore_cap), -zscore_cap)

gene_order_tbl <- shared_gene_table %>%
  filter(gene %in% rownames(heatmap_z)) %>%
  arrange(desc(lps_logFC), masld_padj)

heatmap_z <- heatmap_z[gene_order_tbl$gene, sample_info$sample, drop = FALSE]

annotation_col <- sample_info %>%
  select(sample, group) %>%
  column_to_rownames("sample") %>%
  as.data.frame()

ann_colors <- list(
  group = c(
    "siControl" = "#4D4D4D",
    "siControl + LPS" = "#D55E00",
    "siSLC7A5 + LPS" = "#0072B2"
  )
)

readr::write_csv(
  as.data.frame(heatmap_z) %>% rownames_to_column("gene"),
  file.path(out_dir, "S3A_shared_NFKB_heatmap_zscore_matrix.csv")
)
```

```{r S3A-heatmap, fig.width=5.2, fig.height=6.2}
p_heatmap <- pheatmap(
  heatmap_z,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  annotation_col = annotation_col,
  annotation_colors = ann_colors,
  color = colorRampPalette(c("#2166AC", "white", "#B2182B"))(101),
  breaks = seq(-zscore_cap, zscore_cap, length.out = 102),
  fontsize = 8,
  fontsize_row = ifelse(nrow(heatmap_z) > 35, 5.5, 7),
  fontsize_col = 8,
  border_color = NA,
  main = "Shared NF-kB target genes"
)

pdf(file.path(out_dir, "S3A_shared_NFKB_targets_heatmap.pdf"), width = 5.2, height = 6.2)
print(p_heatmap)
dev.off()

png(file.path(out_dir, "S3A_shared_NFKB_targets_heatmap.png"), width = 5.2, height = 6.2, units = "in", res = 300)
print(p_heatmap)
dev.off()
```

# 8. S3C - NF-kB Target Gene-Set Score

Mean z-score module score is calculated for each sample using all shared NF-kB target genes.

```{r geneset-score-data}
score_df <- heatmap_z %>%
  as.data.frame() %>%
  rownames_to_column("gene") %>%
  pivot_longer(-gene, names_to = "sample", values_to = "zscore") %>%
  group_by(sample) %>%
  summarise(NFKB_target_score = mean(zscore, na.rm = TRUE), .groups = "drop") %>%
  left_join(sample_info, by = "sample")

score_stats <- score_df %>%
  group_by(group) %>%
  summarise(mean_sem(NFKB_target_score), .groups = "drop")

score_pvals <- tibble(
  comparison = c("siControl + LPS vs siControl", "siSLC7A5 + LPS vs siControl + LPS"),
  p_value = c(
    safe_t_test(score_df, "NFKB_target_score", "siControl + LPS", "siControl"),
    safe_t_test(score_df, "NFKB_target_score", "siSLC7A5 + LPS", "siControl + LPS")
  )
) %>%
  mutate(p_adj_BH = p.adjust(p_value, method = "BH"))

readr::write_csv(score_df, file.path(out_dir, "S3C_shared_NFKB_module_score_by_sample.csv"))
readr::write_csv(score_stats, file.path(out_dir, "S3C_shared_NFKB_module_score_group_summary.csv"))
readr::write_csv(score_pvals, file.path(out_dir, "S3C_shared_NFKB_module_score_ttest.csv"))

score_df
score_pvals
```

```{r S3C-geneset-score, fig.width=4.2, fig.height=4.2}
group_colors <- c(
  "siControl" = "#4D4D4D",
  "siControl + LPS" = "#D55E00",
  "siSLC7A5 + LPS" = "#0072B2"
)

p_score <- ggplot(score_df, aes(x = group, y = NFKB_target_score, color = group)) +
  geom_boxplot(width = 0.55, outlier.shape = NA, linewidth = 0.45, color = "grey25") +
  geom_jitter(width = 0.08, size = 2.7, alpha = 0.95) +
  scale_color_manual(values = group_colors) +
  labs(x = NULL, y = "Mean z-score", title = "Shared NF-kB target score") +
  theme_classic(base_size = 11) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 35, hjust = 1),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

p_score

ggsave(file.path(out_dir, "S3C_shared_NFKB_target_score.pdf"), p_score, width = 4.2, height = 4.2)
ggsave(file.path(out_dir, "S3C_shared_NFKB_target_score.png"), p_score, width = 4.2, height = 4.2, dpi = 300)
```

# 9. S3B - Representative NF-kB Target Gene Plots

Representative genes are selected from the candidate list if they are in the shared NF-kB gene set and show the expected bulk RNA-seq pattern. If fewer than 6 candidate genes pass, additional shared NF-kB genes are added based on high LPS induction and stronger reduction by siSLC7A5 knockdown.

```{r representative-gene-selection}
expr_long <- log_expr_mat %>%
  as.data.frame() %>%
  rownames_to_column("gene") %>%
  filter(gene %in% heatmap_genes) %>%
  pivot_longer(-gene, names_to = "sample", values_to = "log2_TPM") %>%
  left_join(sample_info, by = "sample")

gene_pattern <- expr_long %>%
  group_by(gene, group) %>%
  summarise(mean_log2_TPM = mean(log2_TPM), .groups = "drop") %>%
  pivot_wider(names_from = group, values_from = mean_log2_TPM) %>%
  mutate(
    lps_induction = `siControl + LPS` - siControl,
    kd_reduction = `siControl + LPS` - `siSLC7A5 + LPS`,
    expected_pattern = lps_induction > 0 & kd_reduction > 0
  ) %>%
  left_join(shared_gene_table, by = "gene") %>%
  arrange(desc(expected_pattern), desc(kd_reduction), desc(lps_induction), masld_padj)

candidate_selected <- gene_pattern %>%
  filter(gene %in% toupper(representative_candidates), expected_pattern) %>%
  arrange(match(gene, toupper(representative_candidates))) %>%
  pull(gene)

fallback_selected <- gene_pattern %>%
  filter(expected_pattern, !gene %in% candidate_selected) %>%
  pull(gene)

representative_genes <- unique(c(candidate_selected, fallback_selected)) %>%
  head(8)

if (length(representative_genes) < 6) {
  representative_genes <- unique(c(
    representative_genes,
    gene_pattern %>% filter(!gene %in% representative_genes) %>% pull(gene)
  )) %>%
    head(8)
}

readr::write_csv(gene_pattern, file.path(out_dir, "S3B_shared_NFKB_gene_bulk_pattern_summary.csv"))
readr::write_csv(tibble(gene = representative_genes), file.path(out_dir, "S3B_representative_genes_selected.csv"))

representative_genes
```

```{r representative-gene-plot-data}
rep_expr <- expr_long %>%
  filter(gene %in% representative_genes) %>%
  mutate(gene = factor(gene, levels = representative_genes))

rep_stats <- rep_expr %>%
  group_by(gene, group) %>%
  summarise(mean_sem(log2_TPM), .groups = "drop")

rep_pvals <- rep_expr %>%
  group_by(gene) %>%
  group_modify(~ tibble(
    p_LPS_vs_control = safe_t_test(.x, "log2_TPM", "siControl + LPS", "siControl"),
    p_KD_vs_LPS = safe_t_test(.x, "log2_TPM", "siSLC7A5 + LPS", "siControl + LPS")
  )) %>%
  ungroup() %>%
  mutate(
    padj_LPS_vs_control = p.adjust(p_LPS_vs_control, method = "BH"),
    padj_KD_vs_LPS = p.adjust(p_KD_vs_LPS, method = "BH")
  )

rep_sig_anno <- rep_pvals %>%
  select(gene, padj_LPS_vs_control, padj_KD_vs_LPS) %>%
  pivot_longer(
    cols = starts_with("padj_"),
    names_to = "comparison",
    values_to = "padj"
  ) %>%
  mutate(
    gene = factor(gene, levels = representative_genes),
    x_start = ifelse(comparison == "padj_LPS_vs_control", 1, 2),
    x_end = ifelse(comparison == "padj_LPS_vs_control", 2, 3),
    x_mid = (x_start + x_end) / 2,
    label = p_to_star(padj)
  ) %>%
  left_join(
    rep_expr %>%
      group_by(gene) %>%
      summarise(
        y_max = max(log2_TPM, na.rm = TRUE),
        y_min = min(log2_TPM, na.rm = TRUE),
        .groups = "drop"
      ) %>%
      mutate(
        y_range = pmax(y_max - y_min, 0.25),
        y_base = y_max + 0.10 * y_range
      ),
    by = "gene"
  ) %>%
  group_by(gene) %>%
  arrange(comparison, .by_group = TRUE) %>%
  mutate(
    y = y_base + (row_number() - 1) * 0.18 * y_range,
    y_tick = y - 0.04 * y_range
  ) %>%
  ungroup()

readr::write_csv(rep_expr, file.path(out_dir, "S3B_representative_gene_expression_by_sample.csv"))
readr::write_csv(rep_stats, file.path(out_dir, "S3B_representative_gene_expression_group_summary.csv"))
readr::write_csv(rep_pvals, file.path(out_dir, "S3B_representative_gene_ttest.csv"))
readr::write_csv(rep_sig_anno, file.path(out_dir, "S3B_representative_gene_adjusted_pvalue_annotations.csv"))

rep_pvals
```

```{r S3B-representative-gene-plots, fig.width=8.2, fig.height=5.2}
p_rep <- ggplot(rep_expr, aes(x = group, y = log2_TPM, color = group)) +
  geom_boxplot(width = 0.55, outlier.shape = NA, linewidth = 0.35, color = "grey25") +
  geom_jitter(width = 0.08, size = 1.8, alpha = 0.95) +
  geom_segment(
    data = rep_sig_anno,
    aes(x = x_start, xend = x_end, y = y, yend = y),
    inherit.aes = FALSE,
    linewidth = 0.35,
    color = "grey20"
  ) +
  geom_segment(
    data = rep_sig_anno,
    aes(x = x_start, xend = x_start, y = y, yend = y_tick),
    inherit.aes = FALSE,
    linewidth = 0.35,
    color = "grey20"
  ) +
  geom_segment(
    data = rep_sig_anno,
    aes(x = x_end, xend = x_end, y = y, yend = y_tick),
    inherit.aes = FALSE,
    linewidth = 0.35,
    color = "grey20"
  ) +
  geom_text(
    data = rep_sig_anno,
    aes(x = x_mid, y = y + 0.04 * y_range, label = label),
    inherit.aes = FALSE,
    size = 3.2,
    color = "grey10"
  ) +
  facet_wrap(~ gene, scales = "free_y", ncol = 4) +
  scale_color_manual(values = group_colors) +
  labs(x = NULL, y = "log2(TPM + 1)", title = "Representative shared NF-kB targets") +
  theme_classic(base_size = 10) +
  theme(
    legend.position = "none",
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    strip.background = element_rect(fill = "grey92", color = NA),
    strip.text = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

p_rep

ggsave(file.path(out_dir, "S3B_representative_shared_NFKB_targets.pdf"), p_rep, width = 8.2, height = 5.2)
ggsave(file.path(out_dir, "S3B_representative_shared_NFKB_targets.png"), p_rep, width = 8.2, height = 5.2, dpi = 300)
```

# 10. S3B — DESeq2-based Significance

t-test 대신 DESeq2 결과(padj)를 그대로 가져와 유의성 표시를 붙인 버전.
동일한 boxplot + log2(TPM+1) 데이터를 사용하되, 별표만 DESeq2 기준으로 교체합니다.

```{r deseq2-pvals}
kd_deg_file <- slc_input("bulk_nosi_kd_deg")
stopifnot(file.exists(kd_deg_file))

kd_deg <- suppressMessages(readr::read_csv(kd_deg_file)) %>%
  standardize_deg()

# representative_genes 각각에 대해 DESeq2 padj 추출
deseq2_pvals <- tibble(gene = representative_genes) %>%
  left_join(lps_deg %>% select(gene, padj_LPS_vs_control = padj), by = "gene") %>%
  left_join(kd_deg  %>% select(gene, padj_KD_vs_LPS      = padj), by = "gene")

deseq2_pvals
```

```{r deseq2-sig-anno}
rep_sig_anno_deseq2 <- deseq2_pvals %>%
  pivot_longer(
    cols      = starts_with("padj_"),
    names_to  = "comparison",
    values_to = "padj"
  ) %>%
  mutate(
    gene    = factor(gene, levels = representative_genes),
    x_start = ifelse(comparison == "padj_LPS_vs_control", 1, 2),
    x_end   = ifelse(comparison == "padj_LPS_vs_control", 2, 3),
    x_mid   = (x_start + x_end) / 2,
    label   = p_to_star(padj)
  ) %>%
  left_join(
    rep_expr %>%
      group_by(gene) %>%
      summarise(
        y_max = max(log2_TPM, na.rm = TRUE),
        y_min = min(log2_TPM, na.rm = TRUE),
        .groups = "drop"
      ) %>%
      mutate(
        y_range = pmax(y_max - y_min, 0.25),
        y_base  = y_max + 0.10 * y_range
      ),
    by = "gene"
  ) %>%
  group_by(gene) %>%
  arrange(comparison, .by_group = TRUE) %>%
  mutate(
    y      = y_base + (row_number() - 1) * 0.18 * y_range,
    y_tick = y - 0.04 * y_range
  ) %>%
  ungroup()

readr::write_csv(
  deseq2_pvals,
  file.path(out_dir, "S3B_representative_gene_DESeq2_padj.csv")
)
```

```{r S3B-deseq2, fig.width=8.2, fig.height=5.2}
p_rep_deseq2 <- ggplot(rep_expr, aes(x = group, y = log2_TPM, color = group)) +
  geom_boxplot(width = 0.55, outlier.shape = NA, linewidth = 0.35, color = "grey25") +
  geom_jitter(width = 0.08, size = 1.8, alpha = 0.95) +
  geom_segment(
    data = rep_sig_anno_deseq2,
    aes(x = x_start, xend = x_end, y = y, yend = y),
    inherit.aes = FALSE, linewidth = 0.35, color = "grey20"
  ) +
  geom_segment(
    data = rep_sig_anno_deseq2,
    aes(x = x_start, xend = x_start, y = y, yend = y_tick),
    inherit.aes = FALSE, linewidth = 0.35, color = "grey20"
  ) +
  geom_segment(
    data = rep_sig_anno_deseq2,
    aes(x = x_end, xend = x_end, y = y, yend = y_tick),
    inherit.aes = FALSE, linewidth = 0.35, color = "grey20"
  ) +
  geom_text(
    data = rep_sig_anno_deseq2,
    aes(x = x_mid, y = y + 0.04 * y_range, label = label),
    inherit.aes = FALSE, size = 3.2, color = "grey10"
  ) +
  facet_wrap(~ gene, scales = "free_y", ncol = 4) +
  scale_color_manual(values = group_colors) +
  labs(
    x     = NULL,
    y     = "log2(TPM + 1)",
    title = "Representative shared NF-kB targets (DESeq2 significance)"
  ) +
  theme_classic(base_size = 10) +
  theme(
    legend.position  = "none",
    axis.text.x      = element_blank(),
    axis.ticks.x     = element_blank(),
    strip.background = element_rect(fill = "grey92", color = NA),
    strip.text       = element_text(face = "bold"),
    plot.title       = element_text(hjust = 0.5, face = "bold")
  )

p_rep_deseq2

ggsave(file.path(out_dir, "S3B_representative_shared_NFKB_targets_DESeq2.pdf"), p_rep_deseq2, width = 8.2, height = 5.2)
ggsave(file.path(out_dir, "S3B_representative_shared_NFKB_targets_DESeq2.png"), p_rep_deseq2, width = 8.2, height = 5.2, dpi = 300)
```

# 11. Final Output Files

```{r output-files}
list.files(out_dir, full.names = FALSE)
```

# 12. Session Info

```{r session-info}
sessionInfo()
```
