---
title: "Independent bulk RNA-seq cohort validation of SLC7A5-linked genes"
author: "Jaehyun Choi"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r repository-setup, include=FALSE}
.slc_start <- Sys.getenv("SLC7A5_REPO_ROOT", unset = getwd())
.slc_repo <- normalizePath(.slc_start, mustWork = TRUE)
while (!file.exists(file.path(.slc_repo, "R", "paths.R"))) {
  .slc_parent <- dirname(.slc_repo)
  if (identical(.slc_parent, .slc_repo)) stop("Start inside the repository or set SLC7A5_REPO_ROOT.")
  .slc_repo <- .slc_parent
}
source(file.path(.slc_repo, "R", "paths.R"))
slc_setup("hyu_validation")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 6
)
```

# 1. Purpose

This analysis identifies the value for:

> Validation in an independent bulk RNA-seq cohort with fibrosis annotation (n = 159) confirmed that **XX** of these genes, including **SLC7A5**, were consistently elevated in MASLD and positively correlated with fibrosis severity.

Operational definition used here:

1. Candidate genes are genes reduced by siSLC7A5 knockdown under LPS in the prior bulk RNA-seq analysis.
2. A candidate gene is considered validated when it is:
   - elevated in MASLD vs control in the independent HYU bulk cohort, and
   - positively correlated with fibrosis severity.

# 2. Libraries and Parameters

```{r libraries}
library(tidyverse)
library(readxl)
library(limma)
library(pheatmap)
library(RColorBrewer)
library(ggrepel)

filter <- dplyr::filter
select <- dplyr::select
arrange <- dplyr::arrange
```

```{r parameters}
project_dir <- slc_output_dir("03_bulkRNAseq")
data_dir <- file.path(slc_data_root, "03_bulkRNAseq", "01_Data")
script_dir <- file.path(project_dir, "04_Script")
out_dir <- file.path(project_dir, "02_analysis", "HYU_independent_validation")
fig_dir <- file.path(project_dir, "03_Figure", "HYU_independent_validation")

dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)

expected_n <- 159

candidate_kd_file <- slc_input("hyu_candidate_kd")
candidate_lps_file <- slc_input("hyu_candidate_lps")

padj_cutoff_candidate <- 0.05
lfc_cutoff_candidate <- 0

q_cutoff_validation <- 0.05
min_abs_log2fc_validation <- 0
```

# 3. Candidate Gene Set from Previous Bulk RNA-seq

```{r candidate_genes}
res_lps <- read.csv(candidate_lps_file, check.names = FALSE)
res_kd <- read.csv(candidate_kd_file, check.names = FALSE)

candidate_genes <- res_kd %>%
  transmute(
    gene,
    kd_log2FC = log2FoldChange,
    kd_padj = padj
  ) %>%
  filter(
    !is.na(kd_padj),
    kd_padj < padj_cutoff_candidate,
    kd_log2FC < -lfc_cutoff_candidate
  ) %>%
  left_join(
    res_lps %>%
      transmute(
        gene,
        lps_log2FC = log2FoldChange,
        lps_padj = padj
      ),
    by = "gene"
  ) %>%
  select(gene, lps_log2FC, lps_padj, kd_log2FC, kd_padj) %>%
  arrange(kd_padj)

cat("Candidate genes reduced by siSLC7A5 knockdown under LPS:", nrow(candidate_genes), "\n")
cat("SLC7A5 in candidate set:", "SLC7A5" %in% candidate_genes$gene, "\n")

write.csv(
  candidate_genes,
  file.path(out_dir, "candidate_genes_siSLC7A5KD_down.csv"),
  row.names = FALSE
)

head(candidate_genes, 20)
```

# 4. HYU Cohort Data Import

```{r import_hyu}
metadata <- read.csv(
  slc_input("hyu_metadata"),
  fileEncoding = "UTF-8-BOM",
  check.names = FALSE
)

fpkm_raw <- read_excel(slc_input("hyu_fpkm"))

gene_annot <- fpkm_raw %>%
  select(ens_id = 1, gene_symbol = 2) %>%
  mutate(gene_symbol = as.character(gene_symbol))

expr_tbl <- fpkm_raw %>%
  select(-1, -2) %>%
  mutate(across(everything(), as.numeric)) %>%
  bind_cols(gene_symbol = gene_annot$gene_symbol, .) %>%
  filter(!is.na(gene_symbol), gene_symbol != "") %>%
  group_by(gene_symbol) %>%
  summarise(across(everything(), ~ mean(.x, na.rm = TRUE)), .groups = "drop")

expr_fpkm <- expr_tbl %>%
  column_to_rownames("gene_symbol") %>%
  as.matrix()

common_samples <- intersect(colnames(expr_fpkm), metadata$Sample)
expr_fpkm <- expr_fpkm[, common_samples, drop = FALSE]

metadata <- metadata %>%
  filter(Sample %in% common_samples) %>%
  distinct(Sample, .keep_all = TRUE) %>%
  arrange(match(Sample, common_samples))

rownames(metadata) <- metadata$Sample
expr_fpkm <- expr_fpkm[, rownames(metadata), drop = FALSE]

stopifnot(identical(colnames(expr_fpkm), rownames(metadata)))

cat("HYU matched cohort n:", ncol(expr_fpkm), "\n")
cat("One matched sample without fibrosis annotation is excluded below.\n")

table(metadata$Disease, useNA = "ifany")
```

# 5. Metadata Cleaning

```{r metadata_cleaning}
batch_col <- names(metadata)[str_detect(names(metadata), "^기관코드")][1]
stopifnot(!is.na(batch_col))

metadata2 <- metadata %>%
  mutate(
    disease_code = as.character(Disease),
    disease_group = case_when(
      disease_code == "1" ~ "Control",
      disease_code %in% c("2", "3") ~ "MASLD",
      TRUE ~ NA_character_
    ),
    disease_group = factor(disease_group, levels = c("Control", "MASLD")),
    batch = factor(.data[[batch_col]]),
    gender = factor(Gender),
    age = suppressWarnings(as.numeric(Age)),
    fibrosis_raw = as.character(Degree_of_fibrosis),
    fibrosis_stage = case_when(
      fibrosis_raw == "0" ~ 0,
      fibrosis_raw %in% c("1A", "1B", "1C") ~ 1,
      fibrosis_raw == "2" ~ 2,
      fibrosis_raw == "3" ~ 3,
      fibrosis_raw == "4" ~ 4,
      TRUE ~ NA_real_
    )
  ) %>%
  filter(
    !is.na(disease_group),
    !is.na(fibrosis_stage),
    !is.na(batch),
    !is.na(gender),
    !is.na(age)
  )

expr_log <- log2(expr_fpkm[, rownames(metadata2), drop = FALSE] + 1)

cat("Analysis cohort n:", ncol(expr_log), "\n")
if (ncol(expr_log) != expected_n) {
  warning(
    "Analysis cohort n is ", ncol(expr_log),
    ", not expected_n = ", expected_n,
    ". Check fibrosis annotation filtering."
  )
}
cat("Disease group table:\n")
print(table(metadata2$disease_group, useNA = "ifany"))
cat("Fibrosis stage table:\n")
print(table(metadata2$fibrosis_stage, useNA = "ifany"))
```

# 6. MASLD Elevation Test

```{r masld_elevation}
candidate_present <- candidate_genes %>%
  filter(gene %in% rownames(expr_log))

cat("Candidate genes present in HYU FPKM matrix:", nrow(candidate_present), "\n")
cat("SLC7A5 present in HYU FPKM matrix:", "SLC7A5" %in% rownames(expr_log), "\n")

expr_candidate <- expr_log[candidate_present$gene, , drop = FALSE]

test_masld_elevation <- function(gene) {
  x <- as.numeric(expr_log[gene, ])
  group <- metadata2$disease_group

  control_expr <- x[group == "Control"]
  masld_expr <- x[group == "MASLD"]

  tibble(
    gene = gene,
    control_n = length(control_expr),
    masld_n = length(masld_expr),
    control_mean = mean(control_expr, na.rm = TRUE),
    masld_mean = mean(masld_expr, na.rm = TRUE),
    masld_log2FC = masld_mean - control_mean,
    masld_pvalue = wilcox.test(masld_expr, control_expr, exact = FALSE)$p.value
  )
}

elevation_res <- map_dfr(candidate_present$gene, test_masld_elevation) %>%
  mutate(masld_qvalue = p.adjust(masld_pvalue, method = "BH")) %>%
  arrange(masld_qvalue)

write.csv(
  elevation_res,
  file.path(out_dir, "HYU_MASLD_vs_control_candidate_gene_limma.csv"),
  row.names = FALSE
)

head(elevation_res, 20)
```

# 7. Fibrosis Severity Correlation

```{r fibrosis_correlation}
cor_one_gene <- function(gene) {
  md <- metadata2

  x <- as.numeric(expr_log[gene, rownames(md)])
  y <- md$fibrosis_stage

  ok <- is.finite(x) & is.finite(y)
  if (sum(ok) < 3 || length(unique(x[ok])) < 2 || length(unique(y[ok])) < 2) {
    return(tibble(
      gene = gene,
      fibrosis_n = sum(ok),
      fibrosis_rho = NA_real_,
      fibrosis_pvalue = NA_real_
    ))
  }

  test <- suppressWarnings(cor.test(x[ok], y[ok], method = "spearman", exact = FALSE))

  tibble(
    gene = gene,
    fibrosis_n = sum(ok),
    fibrosis_rho = unname(test$estimate),
    fibrosis_pvalue = test$p.value
  )
}

fibrosis_res <- map_dfr(candidate_present$gene, cor_one_gene) %>%
  mutate(fibrosis_qvalue = p.adjust(fibrosis_pvalue, method = "BH")) %>%
  arrange(fibrosis_qvalue)

write.csv(
  fibrosis_res,
  file.path(out_dir, "HYU_candidate_gene_fibrosis_spearman_n159.csv"),
  row.names = FALSE
)

head(fibrosis_res, 20)
```

# 8. Final Validated Gene List and XX

```{r final_validated_genes}
validated_genes <- candidate_present %>%
  select(gene, lps_log2FC, lps_padj, kd_log2FC, kd_padj) %>%
  left_join(elevation_res, by = "gene") %>%
  left_join(fibrosis_res, by = "gene") %>%
  mutate(
    elevated_in_MASLD = !is.na(masld_qvalue) &
      masld_qvalue < q_cutoff_validation &
      masld_log2FC > min_abs_log2fc_validation,
    positive_fibrosis_correlation = !is.na(fibrosis_qvalue) &
      fibrosis_qvalue < q_cutoff_validation &
      fibrosis_rho > 0,
    validated = elevated_in_MASLD & positive_fibrosis_correlation
  ) %>%
  arrange(desc(validated), masld_qvalue, fibrosis_qvalue)

validated_only <- validated_genes %>% filter(validated)
XX <- nrow(validated_only)

cat("XX validated genes:", XX, "\n")
cat("SLC7A5 validated:", "SLC7A5" %in% validated_only$gene, "\n")
cat(
  "Manuscript sentence:\n",
  "Validation in an independent bulk RNA-seq cohort with fibrosis annotation (n = ",
  ncol(expr_log),
  ") confirmed that ",
  XX,
  " of these genes, including SLC7A5, were consistently elevated in MASLD and positively correlated with fibrosis severity.\n",
  sep = ""
)

write.csv(
  validated_genes,
  file.path(out_dir, "HYU_candidate_gene_validation_all_results.csv"),
  row.names = FALSE
)

write.csv(
  validated_only,
  file.path(out_dir, "HYU_validated_genes_for_XX.csv"),
  row.names = FALSE
)

validated_only
```

# 9. Visualization

## 9-1. SLC7A5 Expression and Fibrosis

```{r slc7a5_plots, fig.width=10, fig.height=4}
plot_gene <- "SLC7A5"

if (plot_gene %in% rownames(expr_log)) {
  plot_df <- metadata2 %>%
    mutate(expr = as.numeric(expr_log[plot_gene, rownames(metadata2)]))

  p1 <- ggplot(plot_df, aes(x = disease_group, y = expr, fill = disease_group)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.7) +
    geom_jitter(width = 0.15, size = 1.8, alpha = 0.8) +
    scale_fill_brewer(palette = "Set2") +
    labs(x = NULL, y = "log2(FPKM + 1)", title = paste0(plot_gene, " in MASLD")) +
    theme_bw() +
    theme(legend.position = "none")

  p2 <- ggplot(plot_df, aes(x = fibrosis_stage, y = expr)) +
    geom_point(size = 2, alpha = 0.8) +
    geom_smooth(method = "lm", se = TRUE, color = "#2C7FB8") +
    scale_x_continuous(breaks = sort(unique(plot_df$fibrosis_stage))) +
    labs(x = "Fibrosis stage", y = "log2(FPKM + 1)",
         title = paste0(plot_gene, " vs fibrosis severity")) +
    theme_bw()

  print(p1)
  print(p2)

  ggsave(file.path(fig_dir, "SLC7A5_MASLD_expression_boxplot.png"), p1, width = 5, height = 4, dpi = 300)
  ggsave(file.path(fig_dir, "SLC7A5_fibrosis_correlation.png"), p2, width = 5, height = 4, dpi = 300)
} else {
  cat("SLC7A5 was not found in the expression matrix.\n")
}
```

## 9-2. Validated Gene Heatmap

```{r validated_heatmap, fig.width=10, fig.height=8}
if (nrow(validated_only) >= 2) {
  heat_genes <- validated_only$gene
  heat_mat <- expr_log[heat_genes, , drop = FALSE]

  ann_col <- metadata2 %>%
    select(disease_group, fibrosis_stage, batch) %>%
    as.data.frame()
  rownames(ann_col) <- rownames(metadata2)

  pheatmap(
    heat_mat,
    scale = "row",
    annotation_col = ann_col,
    show_colnames = FALSE,
    clustering_method = "ward.D2",
    color = colorRampPalette(c("#2166AC", "white", "#B2182B"))(200),
    main = paste0("Validated genes in HYU cohort (XX = ", XX, ")")
  )
} else {
  cat("Fewer than two validated genes; heatmap skipped.\n")
}
```

## 9-3. Validation Summary Plot

```{r validation_summary_plot, fig.width=7, fig.height=6}
summary_plot_df <- validated_genes %>%
  mutate(
    label = if_else(validated | gene == "SLC7A5", gene, NA_character_)
  )

p_summary <- ggplot(
  summary_plot_df,
  aes(x = masld_log2FC, y = fibrosis_rho, color = validated, label = label)
) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey60") +
  geom_point(size = 2.4, alpha = 0.85) +
  geom_text_repel(max.overlaps = Inf, size = 3) +
  scale_color_manual(values = c(`FALSE` = "grey65", `TRUE` = "#D95F02")) +
  labs(
    x = "MASLD vs control log2FC",
    y = "Spearman rho with fibrosis stage",
    color = "Validated",
    title = paste0("HYU independent cohort validation (XX = ", XX, ")")
  ) +
  theme_bw()

print(p_summary)
ggsave(file.path(fig_dir, "HYU_validation_summary_scatter.png"), p_summary, width = 7, height = 6, dpi = 300)
```

# 10. Session Info

```{r session_info}
sessionInfo()
```
