---
title: "HYU bulk RNA-seq: SLC7A5 expression by fibrosis severity"
author: "Jaehyun Choi"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r repository-setup, include=FALSE}
.slc_start <- Sys.getenv("SLC7A5_REPO_ROOT", unset = getwd())
.slc_repo <- normalizePath(.slc_start, mustWork = TRUE)
while (!file.exists(file.path(.slc_repo, "R", "paths.R"))) {
  .slc_parent <- dirname(.slc_repo)
  if (identical(.slc_parent, .slc_repo)) stop("Start inside the repository or set SLC7A5_REPO_ROOT.")
  .slc_repo <- .slc_parent
}
source(file.path(.slc_repo, "R", "paths.R"))
slc_setup("hyu_fibrosis")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 6
)
```

# 1. Purpose

This Rmd uses only the independent HYU patient bulk RNA-seq cohort in `01_Data`.
It does not use the siSLC7A5 bulk RNA-seq DEG results.

Main output:

- SLC7A5 expression grouped by fibrosis severity
- SLC7A5 expression grouped by disease category 1/2/3
- SLC7A5 expression grouped by fibrosis category F0/F1/F2-F3
- Summary statistics and association test between SLC7A5 expression and fibrosis severity

# 2. Libraries and Paths

```{r libraries}
library(tidyverse)
library(readxl)
library(RColorBrewer)

filter <- dplyr::filter
select <- dplyr::select
arrange <- dplyr::arrange
```

```{r paths}
project_dir <- slc_output_dir("03_bulkRNAseq")
data_dir <- file.path(slc_data_root, "03_bulkRNAseq", "01_Data")
out_dir <- file.path(project_dir, "02_analysis", "HYU_SLC7A5_fibrosis")
fig_dir <- file.path(project_dir, "03_Figure", "HYU_SLC7A5_fibrosis")

dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)
```

# 3. Import HYU Patient Cohort

```{r import}
metadata <- read.csv(
  slc_input("hyu_metadata"),
  fileEncoding = "UTF-8-BOM",
  check.names = FALSE
)

fpkm_raw <- read_excel(slc_input("hyu_fpkm"))

gene_symbol_col <- names(fpkm_raw)[2]

expr_tbl <- fpkm_raw %>%
  rename(gene_symbol = all_of(gene_symbol_col)) %>%
  select(-ens_id) %>%
  mutate(gene_symbol = as.character(gene_symbol)) %>%
  filter(!is.na(gene_symbol), gene_symbol != "") %>%
  mutate(across(-gene_symbol, as.numeric)) %>%
  group_by(gene_symbol) %>%
  summarise(across(everything(), ~ mean(.x, na.rm = TRUE)), .groups = "drop")

expr_fpkm <- expr_tbl %>%
  column_to_rownames("gene_symbol") %>%
  as.matrix()

common_samples <- intersect(colnames(expr_fpkm), metadata$Sample)

metadata <- metadata %>%
  filter(Sample %in% common_samples) %>%
  distinct(Sample, .keep_all = TRUE) %>%
  arrange(match(Sample, common_samples))

expr_fpkm <- expr_fpkm[, metadata$Sample, drop = FALSE]
stopifnot(identical(colnames(expr_fpkm), metadata$Sample))

cat("Matched HYU patient samples:", ncol(expr_fpkm), "\n")
cat("SLC7A5 found:", "SLC7A5" %in% rownames(expr_fpkm), "\n")

table(metadata$Degree_of_fibrosis, useNA = "ifany")
```

# 4. Clean Fibrosis Severity and SLC7A5 Expression

```{r clean}
stopifnot("SLC7A5" %in% rownames(expr_fpkm))

plot_df <- metadata %>%
  mutate(
    fibrosis_raw = as.character(Degree_of_fibrosis),
    fibrosis_stage = case_when(
      fibrosis_raw == "0" ~ 0,
      fibrosis_raw %in% c("1A", "1B", "1C") ~ 1,
      fibrosis_raw == "2" ~ 2,
      fibrosis_raw == "3" ~ 3,
      fibrosis_raw == "4" ~ 4,
      TRUE ~ NA_real_
    ),
    fibrosis_group = factor(
      paste0("F", fibrosis_stage),
      levels = paste0("F", 0:4)
    ),
    fibrosis_group_3 = case_when(
      fibrosis_stage == 0 ~ "F0",
      fibrosis_stage == 1 ~ "F1",
      fibrosis_stage %in% c(2, 3) ~ "F2/F3",
      fibrosis_stage == 4 ~ "F4",
      TRUE ~ NA_character_
    ),
    fibrosis_group_3 = factor(fibrosis_group_3, levels = c("F0", "F1", "F2/F3", "F4")),
    disease_group = factor(
      as.character(Disease),
      levels = c("1", "2", "3"),
      labels = c("Disease 1", "Disease 2", "Disease 3")
    ),
    slc7a5_fpkm = as.numeric(expr_fpkm["SLC7A5", Sample]),
    slc7a5_log2fpkm = log2(slc7a5_fpkm + 1)
  ) %>%
  filter(!is.na(fibrosis_stage), !is.na(slc7a5_log2fpkm))

cat("Samples used for SLC7A5-fibrosis plot:", nrow(plot_df), "\n")
print(table(plot_df$fibrosis_group, useNA = "ifany"))
print(table(plot_df$fibrosis_group_3, useNA = "ifany"))
print(table(plot_df$disease_group, useNA = "ifany"))

write.csv(
  plot_df,
  file.path(out_dir, "HYU_SLC7A5_expression_by_fibrosis_sample_level.csv"),
  row.names = FALSE
)
```

# 5. Group Summary

```{r summary}
summary_df <- plot_df %>%
  group_by(fibrosis_group, fibrosis_stage) %>%
  summarise(
    n = n(),
    mean_log2fpkm = mean(slc7a5_log2fpkm, na.rm = TRUE),
    sd_log2fpkm = sd(slc7a5_log2fpkm, na.rm = TRUE),
    sem_log2fpkm = sd_log2fpkm / sqrt(n),
    median_log2fpkm = median(slc7a5_log2fpkm, na.rm = TRUE),
    .groups = "drop"
  )

write.csv(
  summary_df,
  file.path(out_dir, "HYU_SLC7A5_expression_by_fibrosis_summary.csv"),
  row.names = FALSE
)

summary_df
```

# 6. Statistics

```{r statistics}
kruskal_res <- kruskal.test(slc7a5_log2fpkm ~ fibrosis_group, data = plot_df)
spearman_res <- cor.test(
  plot_df$slc7a5_log2fpkm,
  plot_df$fibrosis_stage,
  method = "spearman",
  exact = FALSE
)

cat("Kruskal-Wallis p-value:", kruskal_res$p.value, "\n")
cat("Spearman rho:", unname(spearman_res$estimate), "\n")
cat("Spearman p-value:", spearman_res$p.value, "\n")

stats_df <- tibble(
  test = c("Kruskal-Wallis", "Spearman"),
  statistic = c(unname(kruskal_res$statistic), unname(spearman_res$estimate)),
  p_value = c(kruskal_res$p.value, spearman_res$p.value)
)

write.csv(
  stats_df,
  file.path(out_dir, "HYU_SLC7A5_fibrosis_statistics.csv"),
  row.names = FALSE
)
```

# 7. Violin Plot: SLC7A5 by Disease Group

```{r violin_disease, fig.width=7, fig.height=5}
plot_df_disease <- plot_df %>%
  filter(!is.na(disease_group), disease_group %in% c("Disease 1", "Disease 2", "Disease 3")) %>%
  mutate(disease_group = droplevels(disease_group))

disease_kw <- kruskal.test(slc7a5_log2fpkm ~ disease_group, data = plot_df_disease)

disease_pairwise_comparisons <- tribble(
  ~group1,     ~group2,
  "Disease 1", "Disease 2",
  "Disease 2", "Disease 3",
  "Disease 1", "Disease 3"
)

disease_pairwise_stats <- disease_pairwise_comparisons %>%
  mutate(
    p_value = map2_dbl(group1, group2, function(g1, g2) {
      x <- plot_df_disease %>% filter(disease_group == g1) %>% pull(slc7a5_log2fpkm)
      y <- plot_df_disease %>% filter(disease_group == g2) %>% pull(slc7a5_log2fpkm)
      wilcox.test(x, y, exact = FALSE)$p.value
    }),
    p_adj = p.adjust(p_value, method = "BH"),
    p_adj_label = case_when(
      p_adj < 0.0001 ~ "****",
      p_adj < 0.001  ~ "***",
      p_adj < 0.01   ~ "**",
      p_adj < 0.05   ~ "*",
      TRUE           ~ "ns"
    ),
    x = match(group1, levels(plot_df_disease$disease_group)),
    xend = match(group2, levels(plot_df_disease$disease_group))
  )

disease_y_range <- range(plot_df_disease$slc7a5_log2fpkm, na.rm = TRUE)
disease_y_step <- diff(disease_y_range) * 0.12
if (disease_y_step == 0) {
  disease_y_step <- 0.2
}

disease_pairwise_stats <- disease_pairwise_stats %>%
  mutate(
    y = max(plot_df_disease$slc7a5_log2fpkm, na.rm = TRUE) + disease_y_step * row_number(),
    y_tip = y - disease_y_step * 0.18
  )

write.csv(
  disease_pairwise_stats,
  file.path(out_dir, "HYU_SLC7A5_disease_1_2_3_pairwise_wilcox_BH.csv"),
  row.names = FALSE
)

p_violin_disease <- ggplot(plot_df_disease, aes(x = disease_group, y = slc7a5_log2fpkm, fill = disease_group)) +
  geom_violin(trim = FALSE, alpha = 0.75, color = "black", linewidth = 0.25) +
  geom_boxplot(width = 0.16, outlier.shape = NA, alpha = 0.85, color = "black") +
  geom_jitter(width = 0.08, size = 1.7, alpha = 0.7) +
  geom_segment(
    data = disease_pairwise_stats,
    aes(x = x, xend = xend, y = y, yend = y),
    inherit.aes = FALSE,
    linewidth = 0.45
  ) +
  geom_segment(
    data = disease_pairwise_stats,
    aes(x = x, xend = x, y = y, yend = y_tip),
    inherit.aes = FALSE,
    linewidth = 0.45
  ) +
  geom_segment(
    data = disease_pairwise_stats,
    aes(x = xend, xend = xend, y = y, yend = y_tip),
    inherit.aes = FALSE,
    linewidth = 0.45
  ) +
  geom_text(
    data = disease_pairwise_stats,
    aes(x = (x + xend) / 2, y = y + disease_y_step * 0.12, label = p_adj_label),
    inherit.aes = FALSE,
    size = 4.5
  ) +
  scale_fill_brewer(palette = "Set2", drop = FALSE) +
  coord_cartesian(ylim = c(NA, max(disease_pairwise_stats$y) + disease_y_step * 0.45)) +
  labs(
    x = "Disease group",
    y = "SLC7A5 expression (log2 FPKM + 1)",
    title = "SLC7A5 expression by disease group",
    subtitle = paste0("Kruskal-Wallis p = ", signif(disease_kw$p.value, 3))
  ) +
  theme_bw() +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )

print(p_violin_disease)

ggsave(
  file.path(fig_dir, "HYU_SLC7A5_disease_1_2_3_violin.png"),
  p_violin_disease,
  width = 7,
  height = 5,
  dpi = 300
)

ggsave(
  file.path(fig_dir, "HYU_SLC7A5_disease_1_2_3_violin.pdf"),
  p_violin_disease,
  width = 7,
  height = 5,
  dpi = 300
)
```

# 8. Violin Plot: SLC7A5 by Fibrosis Group F0/F1/F2-F3

```{r violin_fibrosis_3group, fig.width=7, fig.height=5}
plot_df_fib3 <- plot_df %>%
  filter(!is.na(fibrosis_group_3), fibrosis_group_3 %in% c("F0", "F1", "F2/F3")) %>%
  mutate(fibrosis_group_3 = droplevels(fibrosis_group_3))

fib3_kw <- kruskal.test(slc7a5_log2fpkm ~ fibrosis_group_3, data = plot_df_fib3)

pairwise_comparisons <- tribble(
  ~group1, ~group2,
  "F0",    "F1",
  "F1",    "F2/F3",
  "F0",    "F2/F3"
)

pairwise_stats <- pairwise_comparisons %>%
  mutate(
    p_value = map2_dbl(group1, group2, function(g1, g2) {
      x <- plot_df_fib3 %>% filter(fibrosis_group_3 == g1) %>% pull(slc7a5_log2fpkm)
      y <- plot_df_fib3 %>% filter(fibrosis_group_3 == g2) %>% pull(slc7a5_log2fpkm)
      wilcox.test(x, y, exact = FALSE)$p.value
    }),
    p_adj = p.adjust(p_value, method = "BH"),
    p_adj_label = case_when(
      p_adj < 0.0001 ~ "****",
      p_adj < 0.001  ~ "***",
      p_adj < 0.01   ~ "**",
      p_adj < 0.05   ~ "*",
      TRUE           ~ "ns"
    ),
    x = match(group1, levels(plot_df_fib3$fibrosis_group_3)),
    xend = match(group2, levels(plot_df_fib3$fibrosis_group_3))
  )

y_range <- range(plot_df_fib3$slc7a5_log2fpkm, na.rm = TRUE)
y_step <- diff(y_range) * 0.12
if (y_step == 0) {
  y_step <- 0.2
}

pairwise_stats <- pairwise_stats %>%
  mutate(
    y = max(plot_df_fib3$slc7a5_log2fpkm, na.rm = TRUE) + y_step * row_number(),
    y_tip = y - y_step * 0.18
  )

write.csv(
  pairwise_stats,
  file.path(out_dir, "HYU_SLC7A5_fibrosis_F0_F1_F2F3_pairwise_wilcox_BH.csv"),
  row.names = FALSE
)

p_violin_fib3 <- ggplot(plot_df_fib3, aes(x = fibrosis_group_3, y = slc7a5_log2fpkm, fill = fibrosis_group_3)) +
  geom_violin(trim = FALSE, alpha = 0.75, color = "black", linewidth = 0.25) +
  geom_boxplot(width = 0.16, outlier.shape = NA, alpha = 0.85, color = "black") +
  geom_jitter(width = 0.08, size = 1.7, alpha = 0.7) +
  geom_segment(
    data = pairwise_stats,
    aes(x = x, xend = xend, y = y, yend = y),
    inherit.aes = FALSE,
    linewidth = 0.45
  ) +
  geom_segment(
    data = pairwise_stats,
    aes(x = x, xend = x, y = y, yend = y_tip),
    inherit.aes = FALSE,
    linewidth = 0.45
  ) +
  geom_segment(
    data = pairwise_stats,
    aes(x = xend, xend = xend, y = y, yend = y_tip),
    inherit.aes = FALSE,
    linewidth = 0.45
  ) +
  geom_text(
    data = pairwise_stats,
    aes(x = (x + xend) / 2, y = y + y_step * 0.12, label = p_adj_label),
    inherit.aes = FALSE,
    size = 4.5
  ) +
  scale_fill_brewer(palette = "Set2", drop = FALSE) +
  coord_cartesian(ylim = c(NA, max(pairwise_stats$y) + y_step * 0.45)) +
  labs(
    x = "Fibrosis group",
    y = "SLC7A5 expression (log2 FPKM + 1)",
    title = "SLC7A5 expression by fibrosis group",
    subtitle = paste0("Kruskal-Wallis p = ", signif(fib3_kw$p.value, 3))
  ) +
  theme_bw() +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )

print(p_violin_fib3)

ggsave(
  file.path(fig_dir, "HYU_SLC7A5_fibrosis_F0_F1_F2F3_violin.png"),
  p_violin_fib3,
  width = 7,
  height = 5,
  dpi = 300
)

ggsave(
  file.path(fig_dir, "HYU_SLC7A5_fibrosis_F0_F1_F2F3_violin.pdf"),
  p_violin_fib3,
  width = 7,
  height = 5,
  dpi = 300
)
```

# 9. Bar Graph: SLC7A5 by Fibrosis Severity

```{r barplot, fig.width=7, fig.height=5}
p_bar <- ggplot(summary_df, aes(x = fibrosis_group, y = mean_log2fpkm, fill = fibrosis_group)) +
  geom_col(width = 0.68, color = "black", linewidth = 0.25) +
  geom_errorbar(
    aes(ymin = mean_log2fpkm - sem_log2fpkm, ymax = mean_log2fpkm + sem_log2fpkm),
    width = 0.18,
    linewidth = 0.45
  ) +
  geom_point(
    data = plot_df,
    aes(x = fibrosis_group, y = slc7a5_log2fpkm),
    inherit.aes = FALSE,
    position = position_jitter(width = 0.12, height = 0),
    size = 1.7,
    alpha = 0.75
  ) +
  geom_text(
    aes(label = paste0("n=", n), y = mean_log2fpkm + sem_log2fpkm + 0.15),
    size = 3.5
  ) +
  scale_fill_brewer(palette = "Set2", drop = FALSE) +
  labs(
    x = "Fibrosis severity",
    y = "SLC7A5 expression (log2 FPKM + 1)",
    title = "SLC7A5 expression by fibrosis severity"
  ) +
  theme_bw() +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5)
  )

print(p_bar)

ggsave(
  file.path(fig_dir, "HYU_SLC7A5_fibrosis_barplot_mean_sem.png"),
  p_bar,
  width = 7,
  height = 5,
  dpi = 300
)
```

# 10. Boxplot Alternative

```{r boxplot, fig.width=7, fig.height=5}
p_box <- ggplot(plot_df, aes(x = fibrosis_group, y = slc7a5_log2fpkm, fill = fibrosis_group)) +
  geom_boxplot(width = 0.62, outlier.shape = NA, alpha = 0.75) +
  geom_jitter(width = 0.12, size = 1.8, alpha = 0.75) +
  scale_fill_brewer(palette = "Set2", drop = FALSE) +
  labs(
    x = "Fibrosis severity",
    y = "SLC7A5 expression (log2 FPKM + 1)",
    title = "SLC7A5 expression by fibrosis severity"
  ) +
  theme_bw() +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5)
  )

print(p_box)

ggsave(
  file.path(fig_dir, "HYU_SLC7A5_fibrosis_boxplot.png"),
  p_box,
  width = 7,
  height = 5,
  dpi = 300
)
```

# 11. Session Info

```{r session_info}
sessionInfo()
```
